---
title: "Compendium of Legislation Applicable to Pacific Salmon and Ecosystems (CoLAPSE)"
author: "Joe Enns, Cory Lagasse"
date: "`r Sys.Date()`"
output:
  bookdown::tufte_html2:
    toc: true
    number_sections: false
    self_contained: true
    fig_caption: true
documentclass: article
link-citations: true
---

<style>
/* Remove bullets and padding from the TOC */
#TOC ul {
  list-style-type: none;
  margin: 0;
  padding-left: 0;
}

/* Adjust line spacing and spacing between items */
#TOC li {
  margin: 0.2em 0;
  line-height: 1.2;
}

/* Style the links */
#TOC a {
  text-decoration: none;
  color: #444;
}

#TOC a:hover {
  color: #005588;
}
</style>


```{r setup, include=FALSE}
library(here)
library(data.table)
library(dplyr)
library(reactable)
library(htmltools)

# Define the file path
rds_path <- here("Full_legislation_compendium.rds")

# Load the full R object
if (file.exists(rds_path)) {
    loaded_data <- readRDS(rds_path)
    message("✅ File loaded successfully!")
} else {
    stop("❌ File does not exist: ", rds_path)
}

# Ensure the dataset is loaded
if (!exists("loaded_data")) {
    stop("❌ Loaded data is missing. Check the file content.")
}

# Extract Full_legislation_parsed_DT from loaded_data
if ("Full_legislation_parsed_DT" %in% names(loaded_data)) {
    Full_legislation_parsed_DT <- loaded_data$Full_legislation_parsed_DT
    message("✅ Extracted 'Full_legislation_parsed_DT' successfully!")
} else {
    stop("❌ Object 'Full_legislation_parsed_DT' not found in the dataset.")
}

# Extract md_threats_keywords from loaded_data
if ("md_threats_keywords" %in% names(loaded_data)) {
    md_threats_keywords <- loaded_data$md_threats_keywords
    message("✅ Extracted 'md_threats_keywords' successfully!")
} else {
    stop("❌ Object 'md_threats_keywords' not found in the dataset.")
}

# Extract clause_type_keywords from loaded_data
if ("clause_type_keywords" %in% names(loaded_data)) {
    clause_type_keywords <- loaded_data$clause_type_keywords
    message("✅ Extracted 'clause_type_keywords' successfully!")
} else {
    stop("❌ Object 'clause_type_keywords' not found in the dataset.")
}

# Ensure objects were properly assigned (but suppress unnecessary printing)
invisible(str(md_threats_keywords))
invisible(str(clause_type_keywords))
```

# INTRODUCTION
## Background
Resource conservation and stewardship in Canada is governed through a decentralized, multi-sector array of constitutional authority administered through federal, provincial, territorial, and municipal governments (Ray et al 2021; Campbell and Thomas 2002). Stemming from the Constitution Act, legislative control over ecological conservation has been a reactionary process of assigning authority to multiple jurisdictions over new environmental issues as they arise, which creates a complex and overlapping distribution of governance (Becklumb 2013). This patchwork of authority is most prominent in Pacific salmon management due to the complicated geographic nature of their anadromous life cycle, spanning a variety of freshwater and marine ecosystems coast-wide and across international boundaries.

While legislation and programs may often operate on a sector-by-sector and individual resource basis, integrated stewardship relies on scale-appropriate planning and decision making (Ray et al 2021). The need for collaborative, transparent decision-making in salmon management is recognized throughout policies, plans, and initiatives, such as the Pacific Salmon Strategy Initiative and the Wild Salmon Policy. Successful conservation and rebuilding of Pacific salmon  therefore requires coordination of activities across discrete administrative responsibilities and regulatory frameworks. However, such coordination requires adequate organizational capacity and systems for effective implementation to ensure that regulatory controls and mitigation activities address underlying threats and factors limiting productivity.

In this report, we present a framework for relating threats to salmon with their legislative and regulatory context within British Columbia. We define discrete areas of administrative responsibility for Pacific salmon as “management domains” and conduct a review of legislation to assess the specificity or scope of each clause in relation to Pacific salmon. Our purpose for creating a management domains framework is to:

<ul>
  <li>facilitate a better understanding of the legislative and regulatory context through which Pacific salmon and their ecosystems are governed in British Columbia;</li>
  <li>relate Pacific salmon threat categories to legislation; and</li>
  <li>support comprehensive management and conservation planning for Pacific salmon by assigning discrete areas for the administration of human activities across jurisdictions.</li>
</ul>

# METHODS
## Defining Management Domains
Management domains represent a reframing of multi-jurisdictional issues by administrative concern rather than specific agency, which can mainstream conservation strategies focused on threats to salmon life history. Since management domains are defined as a “separation of administrative concerns,” it is important that each domain is categorized distinctly to minimize grey areas and blurring between individual domains or the exercise would fail to be useful. Aside from aspects of governance and organizational structure, 13 specific management domains were delineated for the purpose of this exercise. The domains were developed to comprehensively encompass the legislative context, administrative functions, and anthropogenic threats related to Pacific salmon in British Columbia. 

We used the International Union for Conservation of Nature (IUCN) threat categories (Salafsky et al 2008) to classify legislation into anthropogenic activities that affect salmon.  IUCN threat categories are widely used to assess threats as part of IUCN red list assessments, as well as in status assessments for Pacific salmon populations conducted by the Committee for the Status of Endangered Wildlife in Canada (COSEWIC). We assigned level 1 and level 2 threat classification hierarchies from the IUCN categories into management domains to separate anthropogenic activities into different areas of administrative concern. 

```{r table-marginnote, results='asis', echo=FALSE}
# Extract unique management domains, remove NA values, and sort alphabetically
if ("Management Domain" %in% names(Full_legislation_parsed_DT)) {
    unique_management_domains <- sort(na.omit(unique(Full_legislation_parsed_DT$`Management Domain`)))
    
    # Print formatted list as an unordered list (`<ul>`)
    cat("<ul>")
    cat(paste0("<li>", unique_management_domains, "</li>", collapse = "\n"))
    cat("</ul>")
} else {
    stop("❌ The required column 'Management Domain' was not found in the dataset.")
}
```

Historically, much of the management decisions and activities related to salmon have been in response to identifiable threats to salmon health and population status. One common framework used to describe impacts to salmon are the Four H’s: Habitat, Hydropower, Harvest, and Hatcheries. Each management domain, as defined in this report, intentionally aligns with one of the four H’s. However, there are aspects of salmon management associated with governance and agreements rather than one of the other H’s, which was categorized as “Handshaking.” Because the management domains are centered around salmon/habitat impacts, this allows structurally for a clearer alignment with IUCN threat categories.

```{r, echo=FALSE, fig.cap="Flowchart showing alignment of Pacific salmon threat factors (Four H's) with Management Domains in relation to legislation and jurisdiction."}
library(knitr)

# Define the file path using here()
image_path <- here("MGMT-D Flowchart.png")

# Include the image if it exists
if (file.exists(image_path)) {
  knitr::include_graphics(image_path)
} else {
  message("❌ Image file not found: ", image_path)
}
```

## Process
Assessment of legislative context is prone to subjective interpretation. The inclusion and exclusion of both the overall acts and the sections within the acts creates points of assumption and bias. For this exercise we cast as wide a net as possible and created a process that is publicly accessible, automated, and repeatable. Other researchers may repeat this process as future legislation and language changes. The process steps are:

<ol>
<li>Legislation Accumulation and Review (CanLii)</li>
<li>Parsing Legislation by Section and Heading (R)</li>
<li>Assigning Management Domain and IUCN Threats (R)</li>
  <ul>
  <li>Building a Keyword List</li>
  <li>Assigning to Sections/Subsections (R)</li>
  </ul>
<li>Assigning Scope (R)</li>
<li>Assigning Clause Type (R)</li>
</ol>

### 1) Legislation Accumulation and Review
We focused our review on established Canadian federal and British Columbian provincial legislation (acts and regulations) for inclusion in the management domain process. We did not include specific departmental policy frameworks and processes throughout this exercise unless specifically referenced in the statutes, because  these executive strategies are prone to changes in the political, administrative, and budgetary landscape. Indigenous legislation, Yukon Territorial legislation, municipal bylaws, and international conventions were not included explicitly. Although these laws are important for management of Pacific salmon, considering these regionally diverse rules was outside the scope of this legislative review. 

The Canadian Legal Information Institute website, <a href="https://www.canlii.org/">CanLII</a>, provides an excellent resource for searching federal and provincial acts and regulations by keywords. We created a systematic process for querying legislation by:

<ol>
  <li>querying Pacific salmon specific terms (salmon, chinook, sockeye, etc.), which we called "Type A Legislation,"</li>
  <li>extracting adjacent keywords from the Type A Legislation (word frequency analysis),</li>
  <li>querying new keywords and IUCN Threat keywords in CanLII to find "Type B Legislation" (doesn't mention salmon by name),</li>
  <li>filtering legislation regionally (excluding provinces other than BC), and</li>
  <li>downloading HTML files for relevant legislation directly from government websites.</li>
</ol>

Through the CanLII website, the user may look up the keywords queried in each act and regulation chosen. This allows the user to quickly assess the context of the word in the legislation and determine if the section is relevant to Pacific salmon. The full list of legislation reviewed in this process is available in Appendix I. 

### 2) Parsing Legislation from HTML Files
The decentralized overlapping patchwork of Pacific salmon legislation in Canada creates a dynamic such that each act and regulation may contain a variety of administrative responsibilities, possibly attributed to different agencies. However, each act and regulation is comprised of sections and subsections that contain individual clauses that are more specific to a particular subject. For the purpose of the review, we used the individual clause as the sample unit for assignment into management domains and other corresponding attributes.

We parsed the HTML files using R and the <code>rvest</code> package to extract the text of each section and subsection grouped by heading. The text was then cleaned and standardized to remove any extraneous characters or formatting issues. Each section was assigned a unique identifier, and metadata such as the act name, jurisdiction, and legislation type (act or regulation) were recorded.

### 3) Assigning Management Domains and IUCN Threats from Keywords
The Management Domains and IUCN threats were assigned to each section based on a keyword list developed from the IUCN threat categories. The keywords were used to identify specific sections and subsections within the legislation. Each section was then assigned to one or more Management Domain and IUCN threat based on the presence of these keywords.Iteratively, we adjusted the keyword list to ensure that it accurately captured the scope of each Management Domain and IUCN threat. The keyword list is available in Appendix II.

### 4) Assigning Scope
We developed a scope classification system to categorize the sections based on their relevance to Pacific salmon. The scope was determined by the specificity of the language used in the section. The Scope categories were defined as follows:

<ol>
  <li>Salmon - refers to salmon specifically</li>
  <li>Fish - refers to fish that include salmon</li>
  <li>Habitat - refers to any habitat that includes salmon habitat</li>
  <li>Governance - refers to administrative processes and structure</li>
</ol>

Scope 1 was assigned separately to sections with the specific salmon keywords. Scope 2-3 were assigned during the Management Domain and IUCN Threat assignment.  

### 5) Assigning Clause Type from Keywords
Similar the Management Domain assignment, we created a Clause Type keyword list to assess the nature or purposes of the clauses. Using language patterns within the legislation, we grouped Clause Type by:

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Extract unique clause types, remove NA values, and sort alphabetically
if ("Clause_Type" %in% names(clause_type_keywords)) {
    unique_clause_types <- sort(na.omit(unique(clause_type_keywords$Clause_Type)))
    
    # Print formatted list as an unordered list (`<ul>`)
    cat("<ul>")
    cat(paste0("<li>", unique_clause_types, "</li>", collapse = "\n"))
    cat("</ul>")
} else {
    stop("❌ The required column 'Clause_Type' was not found in the dataset.")
}
```

# RESULTS
## 1) Compendium of Legislation Applicable to Pacific Salmon and Ecosystems (CoLAPSE)
A summary of results from the initial iterations of the automated compendium are shown below. As the keywords become less specific to salmon, the number of Procedural Elements increases. The compendium is available as an R object, which can be used to query specific sections and subsections by Management Domain, IUCN Threat, Scope, and Clause Type. The full compendium is available in the <code>Full_legislation_compendium.rds</code> file and can be loaded into R using the <code>readRDS()</code> function. The compendium is also available as a CSV file for download from GitHub.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)

# Check if required columns exist
if (all(c("Scope", "Paragraph") %in% names(Full_legislation_parsed_DT))) {
    
    # Remove NA values and blank Paragraphs, then count occurrences by Scope
    scope_count <- Full_legislation_parsed_DT %>%
        filter(!is.na(Scope) & !is.na(Paragraph) & Paragraph != "") %>%
        group_by(Scope) %>%
        summarise(count = n()) %>%
        arrange(Scope)  # Ordered by Scope alphabetically

    # Print formatted output without bullets
    cat("<h3>Procedural Element Count by Scope</h3>")
    cat("<ul style='list-style-type: none; padding: 0; margin: 0;'>")
    for (i in seq_len(nrow(scope_count))) {
        cat(paste0("<li><strong>", scope_count$Scope[i], ":</strong> ", scope_count$count[i], " Procedural Elements</li>"))
    }
    cat("</ul>")
    
} else {
    stop("❌ The required columns 'Scope' or 'Paragraph' were not found in the dataset.")
}
```

Using the '1 - Salmon' scope for Procedural Elements shows the Type A Legislation along with a comparison of the number of Procedural Elements by Jurisdiction. The results are shown in the bar chart below, which displays the number of Procedural Elements for each piece of legislation that is specific to salmon. The chart is color-coded by jurisdiction (Federal or Provincial) to highlight the distribution of Procedural Elements across different legislative frameworks.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
library(dplyr)
library(ggplot2)

# Verify required columns exist
if (all(c("Legislation Name", "Paragraph", "Scope", "Jurisdiction") %in% names(Full_legislation_parsed_DT))) {

    # Filter data where Scope is '1 - Salmon' and count Paragraphs by Legislation Name
    summary_data <- Full_legislation_parsed_DT %>%
        filter(Scope == "1 - Salmon" & !is.na(Paragraph)) %>%
        group_by(`Legislation Name`, Jurisdiction) %>%
        summarise(Paragraph_Count = n()) %>%
        arrange(desc(Paragraph_Count))

    # Ensure data is not empty before plotting
    if (nrow(summary_data) > 0) {
        # Create the bar chart with Jurisdiction color coding
        ggplot(summary_data, aes(x = reorder(`Legislation Name`, Paragraph_Count), y = Paragraph_Count, fill = Jurisdiction)) +
    geom_bar(stat = "identity", color = "black") +
    coord_flip() +  # Horizontal bar chart for readability
    scale_fill_manual(values = c("Federal" = "#8B0000", "Provincial" = "#00008B")) +  # Custom colors
    labs(
        title = "Procedural Element Count by Legislation (Scope: 1 - Salmon)",
        x = "Legislation",
        y = "Procedural Element Count",
        fill = "Jurisdiction"
    ) +
    theme_minimal()

    } else {
        warning("⚠️ No data available for Scope '1 - Salmon'. Check the Scope values in your dataset.")
    }

} else {
    stop("❌ The required columns 'Legislation Name', 'Paragraph', 'Scope', or 'Jurisdiction' were not found in the dataset.")
}
```

## Summary of Management Domains

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
library(dplyr)
library(ggplot2)

# Prepare data: Count distinct Legislation Names for each Management Domain, grouped by Jurisdiction
domain_counts <- Full_legislation_parsed_DT %>%
    filter(!is.na(`Management Domain`) & !is.na(`Legislation Name`) & !is.na(Jurisdiction)) %>%
    distinct(`Management Domain`, `Legislation Name`, Jurisdiction) %>%
    group_by(`Management Domain`, Jurisdiction) %>%
    summarise(Legislation_Count = n(), .groups = "drop") %>%
    arrange(desc(Legislation_Count))  # Sort by count

# Generate stacked bar chart split by Jurisdiction
ggplot(domain_counts, aes(x = reorder(`Management Domain`, Legislation_Count), y = Legislation_Count, fill = Jurisdiction)) +
    geom_bar(stat = "identity", color = "black", width = 0.9) +
    coord_flip() +  # Horizontal bars for readability
    scale_fill_manual(values = c("Federal" = "#8B0000", "Provincial" = "#00008B")) +  # Assign colors
    labs(
        title = "Number of Legislation Names per Management Domain (Split by Jurisdiction)",
        x = "Management Domain",
        y = "Legislation Count",
        fill = "Jurisdiction"
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 12))
```

## Summary of IUCN Threats

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
library(dplyr)
library(ggplot2)

# Prepare data: Count distinct Legislation Names for each L2, grouped by Jurisdiction
domain_counts <- Full_legislation_parsed_DT %>%
    filter(!is.na(L2) & !is.na(`Legislation Name`) & !is.na(Jurisdiction)) %>%
    distinct(L2, `Legislation Name`, Jurisdiction) %>%
    group_by(L2, Jurisdiction) %>%
    summarise(Legislation_Count = n(), .groups = "drop") %>%
    arrange(desc(Legislation_Count))  # Sort by count

# Generate stacked bar chart split by Jurisdiction
ggplot(domain_counts, aes(x = reorder(L2, Legislation_Count), y = Legislation_Count, fill = Jurisdiction)) +
    geom_bar(stat = "identity", color = "black", width = 0.9) +
    coord_flip() +  # Horizontal bars for readability
    scale_fill_manual(values = c("Federal" = "#8B0000", "Provincial" = "#00008B")) +  # Assign colors
    labs(
        title = "Count of Legislation per IUCN Threat (Split by Jurisdiction)",
        x = "IUCN Threat",
        y = "Legislation Count",
        fill = "Jurisdiction"
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 12))
```

## Summary of Clause Types
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
library(dplyr)
library(ggplot2)

# Prepare data: Count Paragraphs per Clause Type while excluding NA Scope rows
clause_data <- Full_legislation_parsed_DT %>%
    filter(!is.na(Clause_Type) & !is.na(Paragraph) & !is.na(Scope) & !is.na(`Legislation Name`)) %>%
    distinct(Clause_Type, Paragraph, `Legislation Name`) %>%
    group_by(Clause_Type) %>%
    summarise(
        Paragraph_Count = n(),
        Legislation_Count = n_distinct(`Legislation Name`),
        .groups = "drop"
    ) %>%
    arrange(desc(Paragraph_Count))  # Sort by Paragraph count

# Generate dot plot
ggplot(clause_data, aes(x = Clause_Type, y = Paragraph_Count)) +
    geom_point(aes(size = Legislation_Count), color = "steelblue", alpha = 0.7) +  # Dot size reflects Legislation Count
    scale_size_continuous(range = c(3, 10)) +  # Adjust dot size range for clarity
    labs(
        title = "Paragraph Count by Clause Type (Dot Size = Number of Legislation Names)",
        x = "Clause Type",
        y = "Paragraph Count",
        size = "Legislation Count"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

<hr>

## 2) Interactive Dashboard

<hr>


# REFERENCES
<style>
.references {
  padding-left: 40px;
  text-indent: -40px;
}
</style>
<div class="references">
Armitage, D., de Loë, R., Plummer, R., 2012. Environmental governance and its implications for conservation practice. Conserv. Lett. 5, 245–255.

Beazley, K.F., and Olive, A. 2021. Transforming conservation in Canada: shifting policies and paradigms. FACETS, 6: 1714–1727. doi:10.1139/ facets-2021-0144.

Becklumb, P. 2013. Federal and provincial jurisdiction to regulate environmental issues. Ottawa.

Bennett NJ, Roth R, Klain SC, Chan K, Christie P, Clark DA, et al. 2017a. Conservation social science: understanding and integrating human dimensions to improve conservation. Biological Conservation, 205: 93–108. DOI: 10.1016/j.biocon.2016.10.006

Campbell ML, and Thomas VG. 2002. Constitutional impacts on conservation—effects of federalism on biodiversity protection. Environment Policy and Law, 32: 223–233.

Collins L, and Sossin L. 2019. Approach to constitutional principles and environmental discretion in Canada. UBC Law Review, 52(1): 293–343. [online]: Available from digitalcommons.osgoode.yorku.ca

Dufresne Y, and Ouellet C. 2019. Conceptualization, measurement and effects of positional issues in the Canadian electoral context. Heliyon, 5(4): e01453. PMID:31025009. DOI:10.1016/j.heliyon. 2019.e01453

Kraus, D., Murphy, S., and Armitage, D. 2021. Ten bridges on the road to recovering Canada’s endangered species. doi:10.1139/facets.

Hutchings JA, Côté IM, Dodson JJ, Fleming IA, Jennings S, Mantua NJ, et al. 2012. Is Canada fulfilling its obligations to sustain marine biodiversity? A summary review, conclusions, and recommendations. Environmental Reviews 20: 353–361. DOI: 10.1139/er-2012-0049

Marentette, J.R. and Kronlund, A.R. 2020. A Cross-Jurisdictional Review of International Fisheries Policies, Standards and Guidelines: Considerations for a Canadian Science Sector Approach. Can. Tech. Rep. Fish. Aquat. Sci. 3342: xiii + 169 p.

McCune JL, Harrower WL, Avery-Gomm S, Brogan JM, Csergő A-M, Davidson LNK, et al. 2013. Threats to Canadian species at risk: an analysis of finalized recovery strategies. Biological Conservation, 166: 254–265. DOI: 10.1016/j.biocon.2013.07.006

Ray, J.C., Grimm, J., and Olive, A. 2021. The biodiversity crisis in Canada: failures and challenges of federal and sub-national strategic and legal frameworks. FACETS. 6: 1044–1068. doi:10.1139/facets.

Salafsky, N., Salzer, D., Stattersfield, A.J., Hilton-Taylor, C., Nuegarten, R., Butchart, S.H.M., Collen, B., Cox, N., Master, L.L., O’Connor, S., Wilkie, D., 2008. A standard lexicon for biodiversity conservation: unified classifications of threats and actions. Conserv. Biol. 22, 897–911.

Swerdfager, Trevor, and Andrea Olive. “Laws Matter: A Foundational Approach to Biodiversity Conservation in Canada.” Edited by David Lesbarreres. FACETS 8 (January 1, 2023): 1–13.

Westwood AR, Otto SP, Mooers A, Darimont C, Hodges KE, Johnson C, et al. 2019. Protecting biodiversity in British Columbia: recommendations for developing species at risk legislation. FACETS, 4(1): 136–160. DOI: 10.1139/facets-2018-0042
</div>

<hr>

# APPENDICES

<hr>

# Appendix I - Legislation Accumulation and Review

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}


# Check if required columns exist
if (all(c("Jurisdiction", "Legislation Name", "Legislation Type") %in% names(Full_legislation_parsed_DT))) {
    
    # Count Acts by Jurisdiction
    acts_count <- Full_legislation_parsed_DT %>%
        filter(`Legislation Type` == "Act") %>%
        group_by(Jurisdiction) %>%
        summarise(count = n_distinct(`Legislation Name`)) %>%
        arrange(desc(count))
    
    # Count Regulations by Jurisdiction
    regs_count <- Full_legislation_parsed_DT %>%
        filter(`Legislation Type` == "Regulations") %>%
        group_by(Jurisdiction) %>%
        summarise(count = n_distinct(`Legislation Name`)) %>%
        arrange(desc(count))
    
    # Print formatted output
    cat("<h2>Acts per Jurisdiction</h2><ul>")
    for (i in seq_len(nrow(acts_count))) {
        cat(paste0("<li><strong>", acts_count$Jurisdiction[i], ":</strong> ", acts_count$count[i], " Acts</li>"))
    }
    cat("</ul>")
    
    cat("<h2>Regulations per Jurisdiction</h2><ul>")
    for (i in seq_len(nrow(regs_count))) {
        cat(paste0("<li><strong>", regs_count$Jurisdiction[i], ":</strong> ", regs_count$count[i], " Regulations</li>"))
    }
    cat("</ul>")
    
} else {
    stop("❌ The required columns 'Jurisdiction', 'Legislation Name', or 'Legislation Type' were not found in the dataset.")
}
```

<hr>

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)
library(tidyr)

# Check if required columns exist
if (all(c("Jurisdiction", "Act Name", "Legislation Name", "Legislation Type") %in% names(Full_legislation_parsed_DT))) {
    
    # Get all distinct Acts grouped by Jurisdiction
    all_acts <- Full_legislation_parsed_DT %>%
        filter(`Legislation Type` == "Act") %>%
        distinct(Jurisdiction, `Act Name`) %>%
        arrange(Jurisdiction, `Act Name`)  # Sorting jurisdictions properly
    
    # Get Acts with Regulations
    acts_with_regs <- Full_legislation_parsed_DT %>%
        filter(`Legislation Type` == "Regulations") %>%
        group_by(Jurisdiction, `Act Name`) %>%
        summarise(regs_list = paste(sort(unique(`Legislation Name`)), collapse="<br>")) %>%
        arrange(Jurisdiction, `Act Name`)
    
    # Ensure all Acts are included, even if they have no Regulations
    full_list <- left_join(all_acts, acts_with_regs, by = c("Jurisdiction", "Act Name")) %>%
        mutate(regs_list = replace_na(regs_list, "No regulations included"))
    
    # Print formatted output grouped by sorted Jurisdiction
    cat("<h2>Acts and Their Regulations by Jurisdiction</h2>")

    jurisdictions <- sort(unique(full_list$Jurisdiction))  # Now properly sorted

    for (j in jurisdictions) {
        cat(paste0("<p style='font-size:18pt; font-weight:bold; margin:10px 0; padding:5px;'>", j, "</p>"))  # Larger jurisdiction headers
        subset <- full_list %>% filter(Jurisdiction == j)
        for (i in seq_len(nrow(subset))) {
            cat(paste0("<p style='font-size:12pt; line-height:1.4; margin:5px 0;'>",
                       "<strong><u>", subset$`Act Name`[i], "</u></strong><br>", subset$regs_list[i], "</p>"))  # Underlined Act Name
        }
    }
    
} else {
    stop("❌ The required columns 'Jurisdiction', 'Act Name', 'Legislation Name', or 'Legislation Type' were not found in the dataset.")
}
```
<hr>
# Appendix II - Management Domain Keyword List
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)

# Check if required columns exist
if (all(c("L1", "L2", "Management Domain") %in% names(md_threats_keywords))) {

    # Remove NA values and sort data
    sorted_keywords <- md_threats_keywords %>%
        filter(!is.na(L1) & !is.na(L2) & !is.na(`Management Domain`)) %>%
        distinct(`Management Domain`, L1, L2) %>%  # Remove duplicates
        arrange(`Management Domain`, L1, L2)

    # Print formatted output grouped by Management Domain
    cat("<h2 style='font-size:22pt; margin-bottom:10px;'>Management Domains & IUCN Threats</h2>")

    for (domain in unique(sorted_keywords$`Management Domain`)) {
        cat(paste0("<h3 style='font-size:20pt; margin-top:12px; padding:0;'>", domain, "</h3>"))
        domain_subset <- sorted_keywords %>% filter(`Management Domain` == domain)

        for (i in seq_len(nrow(domain_subset))) {
            cat(paste0("<p style='font-size:12pt; margin:2px 0; line-height:1.2;'>",
                       "<strong>", domain_subset$L1[i], ":</strong> ", domain_subset$L2[i], "</p>"))
        }
    }

} else {
    stop("❌ The required columns 'L1', 'L2', or 'Management Domain' were not found in the dataset.")
}
```

<hr>

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)

# Check if required columns exist
if (all(c("Management Domain", "Keyword") %in% names(md_threats_keywords))) {
    
    # Remove NA values and sort data correctly
    sorted_keywords <- md_threats_keywords %>%
        filter(!is.na(Keyword) & Keyword != "") %>%
        arrange(`Management Domain`, Keyword)

    # Print formatted output grouped by Management Domain
    cat("<h2 style='font-size:22pt; margin-bottom:10px;'>Keywords by Management Domain</h2>")

    for (domain in unique(sorted_keywords$`Management Domain`)) {
        cat(paste0("<h3 style='font-size:20pt; margin-top:15px;'>", domain, "</h3>"))
        
        domain_subset <- sorted_keywords %>% filter(`Management Domain` == domain)

        # Create a two-column layout
        cat("<div style='display: grid; grid-template-columns: 0.1fr 0.1fr; gap:0px;'>")
        for (keyword in domain_subset$Keyword) {
            cat(paste0("<div style='font-size:11pt; padding:2px;'>", keyword, "</div>"))
        }
        cat("</div>")
    }

} else {
    stop("❌ The required columns 'Management Domain' or 'Keyword' were not found in the dataset.")
}
```

<hr>


# Appendix III - Clause Type Keywords
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)

# Check if required columns exist
if (all(c("Clause_Type", "Keyword") %in% names(clause_type_keywords))) {
    
    # Remove NA values and sort data correctly
    sorted_keywords <- clause_type_keywords %>%
        filter(!is.na(Keyword)) %>%
        arrange(Clause_Type, Keyword)
    
    # Print formatted output grouped by Clause_Type
    cat("<h2 style='font-size:22pt; margin-bottom:10px;'>Clause Type Keywords</h2>")
    
    for (clause_type in unique(sorted_keywords$Clause_Type)) {
        cat(paste0("<h3 style='font-size:20pt; margin-top:15px;'>", clause_type, "</h3>"))
        clause_subset <- sorted_keywords %>% filter(Clause_Type == clause_type)
        
        # Print keywords in a two-column grid layout without bullets
        cat("<div style='display: grid; grid-template-columns: 0.1fr 0.1fr; gap:0px;'>")
        for (keyword in clause_subset$Keyword) {
            cat(paste0("<div style='font-size:11pt; padding:2px;'>", keyword, "</div>"))
        }
        cat("</div>")
    }
    
} else {
    stop("❌ The required columns 'Clause_Type' or 'Keyword' were not found in the dataset.")
}
```

<hr>

# Appendix IV - Example: The Fisheries Act
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)

# Check if required columns exist
if (all(c("Scope", "Section", "Heading", "Legislation Name") %in% names(Full_legislation_parsed_DT))) {
    
    # Filter dataset for 'Fisheries Act (R.S.C., 1985, c. F-14)' and remove NA Scope values, sorting by Section
    filtered_sections <- Full_legislation_parsed_DT %>%
        filter(`Legislation Name` == "Fisheries Act (R.S.C., 1985, c. F-14)" & !is.na(Scope)) %>%
        select(Scope, Section, Heading) %>%
        arrange(Scope, as.numeric(Section))  # Ensures Sections are sorted numerically

   filtered_sections <- Full_legislation_parsed_DT %>%
    filter(`Legislation Name` == "Fisheries Act (R.S.C., 1985, c. F-14)" & !is.na(Scope)) %>%
    select(Scope, Section, Heading) %>%
    distinct() %>%  # Removes duplicate Section-Heading pairs
    arrange(Scope, as.numeric(Section))  # Sorts numerically
    
     # Print formatted output grouped by Scope
    cat("<h2 style='font-size:22pt; margin-bottom:10px;'>Sections and Headings in Fisheries Act</h2>")

    for (scope_val in unique(filtered_sections$Scope)) {
        cat(paste0("<h3 style='font-size:20pt; margin-top:15px;'>", scope_val, "</h3>"))
        
        scope_subset <- filtered_sections %>% filter(Scope == scope_val)

        for (i in 1:nrow(scope_subset)) {
            cat(paste0("<p style='font-size:12pt;margin:2px 0; line-height:1.2;'>", scope_subset$Section[i], " - ", scope_subset$Heading[i], "</p>"))
        }
    }

} else {
    stop("❌ The required columns 'Scope', 'Section', 'Heading', or 'Legislation Name' were not found in the dataset.")
}
```

<hr>

# Appendix V - Example: The Water Sustainability Act
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(dplyr)

# Check if required columns exist
if (all(c("Scope", "Section", "Heading", "Legislation Name") %in% names(Full_legislation_parsed_DT))) {
    
    # Filter dataset for 'Water Sustainability Act' and remove NA Scope values, sorting by Section
    filtered_sections <- Full_legislation_parsed_DT %>%
        filter(`Legislation Name` == "Water Sustainability Act" & !is.na(Scope)) %>%
        select(Scope, Section, Heading) %>%
        arrange(Scope, as.numeric(Section))  # Ensures Sections are sorted numerically

    filtered_sections <- Full_legislation_parsed_DT %>%
    filter(`Legislation Name` == "Water Sustainability Act" & !is.na(Scope)) %>%
    select(Scope, Section, Heading) %>%
    distinct() %>%  # Removes duplicate Section-Heading pairs
    arrange(Scope, as.numeric(Section))  # Sorts numerically

    # Print formatted output grouped by Scope
    cat("<h2 style='font-size:22pt; margin-bottom:10px;'>Sections and Headings in Water Sustainability Act</h2>")

    for (scope_val in unique(filtered_sections$Scope)) {
        cat(paste0("<h3 style='font-size:20pt; margin-top:15px;'>", scope_val, "</h3>"))
        
        scope_subset <- filtered_sections %>% filter(Scope == scope_val)

        for (i in 1:nrow(scope_subset)) {
            cat(paste0("<p style='font-size:12pt; margin:2px 0; line-height:1.2;'>", scope_subset$Section[i], " - ", scope_subset$Heading[i], "</p>"))
        }
    }

} else {
    stop("❌ The required columns 'Scope', 'Section', 'Heading', or 'Legislation Name' were not found in the dataset.")
}
```


# Appendix VI - Compendium Viewer

The following table contains the full database of legislation and the classification of clauses by IUCN threat level 2 (L2), scope, and clause type.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(dplyr)
library(reactable)
library(htmltools)

# Step 1: Prepare cleaned base data
nested_data <- Full_legislation_parsed_DT %>%
  filter(
    !is.na(`Management Domain`), !is.na(L1), !is.na(L2),
    !is.na(Jurisdiction), !is.na(`Legislation Name`),
    !is.na(Section), !is.na(Heading), !is.na(Paragraph)
  ) %>%
  distinct(`Management Domain`, L1, L2, Jurisdiction,
           `Legislation Name`, Heading, Section, Paragraph)

# Step 2: Precompute Paragraphs preserving their original row order
paragraph_lookup <- nested_data %>%
  group_by(`Legislation Name`, Heading, Section) %>%
  summarise(
    Paragraphs = toString(Paragraph),  # preserves input order
    .groups = "drop"
  )

# Step 3: Construct nested explorer structure
l1_nested <- nested_data %>%
  group_by(`Management Domain`, L1, L2, Jurisdiction, `Legislation Name`) %>%
  summarise(Section_data = list(cur_data()), .groups = "drop") %>%
  group_by(`Management Domain`, L1, L2, Jurisdiction) %>%
  summarise(Legislation_data = list(cur_data()), .groups = "drop") %>%
  group_by(`Management Domain`, L1, L2) %>%
  summarise(Jurisdiction_data = list(cur_data()), .groups = "drop") %>%
  group_by(`Management Domain`, L1) %>%
  summarise(L2_data = list(cur_data()), .groups = "drop")

domain_summary <- l1_nested %>%
  group_by(`Management Domain`) %>%
  summarise(L1_data = list(cur_data()), .groups = "drop") %>%
  arrange(`Management Domain`)

# Step 4: White arrow CSS injection
white_arrow_css <- tags$head(
  tags$style(HTML("
    .reactable .rt-expander {
      color: white !important;
    }
    .reactable .rt-expander:focus {
      outline: none;
    }
  "))
)

# Step 5: Row styling for nested dark levels
dark_row <- function(index) {
  list(background = "darkgrey", color = "white")
}

# Step 6: Render final nested explorer
htmltools::browsable(
  tagList(
    white_arrow_css,
    reactable(
      domain_summary,
      columns = list(
        `Management Domain` = colDef(name = "Management Domain"),
        L1_data = colDef(show = FALSE)
      ),
      searchable = TRUE,
      bordered = TRUE,
      striped = TRUE,
      highlight = TRUE,
      defaultPageSize = 10,
      details = function(domain_index) {
        l1_table <- domain_summary$L1_data[[domain_index]]

        reactable(
          l1_table,
          rowStyle = dark_row,
          columns = list(
            L1 = colDef(name = "IUCN Threat Level 1"),
            L2_data = colDef(show = FALSE)
          ),
          details = function(l1_index) {
            l2_table <- l1_table$L2_data[[l1_index]]

            reactable(
              l2_table,
              rowStyle = dark_row,
              columns = list(
                L2 = colDef(name = "IUCN Threat Level 2"),
                Jurisdiction_data = colDef(show = FALSE)
              ),
              details = function(l2_index) {
                jurisdiction_table <- l2_table$Jurisdiction_data[[l2_index]]

                reactable(
                  jurisdiction_table,
                  rowStyle = dark_row,
                  columns = list(
                    Jurisdiction = colDef(name = "Jurisdiction"),
                    Legislation_data = colDef(show = FALSE)
                  ),
                  details = function(j_index) {
                    legislation_table <- jurisdiction_table$Legislation_data[[j_index]]

                    reactable(
                      legislation_table,
                      rowStyle = dark_row,
                      columns = list(
                        `Legislation Name` = colDef(name = "Legislation"),
                        Section_data = colDef(show = FALSE)
                      ),
                      details = function(law_index) {
                        leg_name <- legislation_table$`Legislation Name`[[law_index]]
                        section_data <- legislation_table$Section_data[[law_index]]

                        section_table <- section_data %>%
                          distinct(Heading, Section) %>%
                          mutate(`Legislation Name` = leg_name)

                        table_data <- section_table %>%
                          left_join(paragraph_lookup,
                                    by = c("Legislation Name", "Heading", "Section")) %>%
                          select(Heading, Section, Paragraphs) %>%
                          arrange(Heading, Section)

                        reactable(
                          table_data,
                          columns = list(
                            Heading = colDef(name = "Heading"),
                            Section = colDef(name = "Section"),
                            Paragraphs = colDef(name = "Paragraph(s)")
                          ),
                          compact = TRUE,
                          highlight = TRUE,
                          bordered = TRUE,
                          striped = TRUE,
                          defaultPageSize = 5
                        )
                      }
                    )
                  }
                )
              }
            )
          }
        )
      }
    )
  )
)


```

