---
title: "Compendium of Legislation Applicable to Pacific Salmon and Ecosystems (CoLAPSE)"
author: "Joe Enns, Cory Lagasse"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    self_contained: true
    fig_caption: true
    code_folding: none
    css: "style.css"
    theme: flatly
    highlight: tango
---

# INTRODUCTION

## Background

Resource conservation and stewardship in Canada is governed through a decentralized, multi-sector array of constitutional authority administered through federal, provincial, territorial, and municipal governments (Ray et al 2021; Campbell and Thomas 2002). Stemming from the Constitution Act, legislative control over ecological conservation has been a reactionary process of assigning authority to multiple jurisdictions over new environmental issues as they arise, which creates a complex and overlapping distribution of governance (Becklumb 2013). This patchwork of authority is most prominent in Pacific salmon management due to the complicated geographic nature of their anadromous life cycle, spanning a variety of freshwater and marine ecosystems coast-wide and across international boundaries.

While legislation and programs may often operate on a sector-by-sector and individual resource basis, integrated stewardship relies on scale-appropriate planning and decision making (Ray et al 2021). The need for collaborative, transparent decision-making in salmon management is recognized throughout policies, plans, and initiatives, such as the Pacific Salmon Strategy Initiative and the Wild Salmon Policy. Successful conservation and rebuilding of Pacific salmon populations therefore requires coordination of activities across discrete administrative responsibilities and regulatory frameworks. However, such coordination requires adequate organizational capacity and systems for effective implementation to ensure that regulatory controls and mitigation activities address underlying threats and factors that limit capacity and reduce productivity.

<br>

## Purpose and Objectives

This report outlines a framework and process to relate human activities that threaten Pacific salmon stocks, biology, and ecology to their regulatory statutes. The overall goal of this effort is to better understand the legislative context that governs Pacific salmon in British Columbia related to human activities by administrative concern in order to support conservation planning across jurisdictions. To achieve this, the more specific objectives are to:

<div style="margin-left: 30px; padding-left: 10px">
<ol>
<li>create a process to categorize specific legislation by anthropogenic threats to Pacific salmon;</li>
<li>assess legislative factors by threat categorization; and</li>
<li>connect studied biological and ecological concerns in a current salmon management context.</li>
</ol>
</div>

<br>

# METHODS

## Legislation by Anthrogogenic Threats to Salmon

Legislation governs human activity, which in turn impacts Pacific salmon population and habitat. Federal and provincial acts and regulations in British Columbia might be designed to address one broad category of human activity; however, this activity can cover many aspects of salmon management. For example, the provincial Forest and Range Practices Act is specific to forestry practices and planning, along with cattle ranging practices, but in terms of salmon threats, these activities impact many aspects including stream hydrology, water temperature, effluent, riparian habitat destruction, watercourse changes from road infrastructure, and the spread of invasive species. To better categorize legislation aligned with salmon threats and create a “buffer” between biological management and the regulatory context, this study defines discrete areas of administrative responsibility for Pacific salmon as “management domains.”

<br>

### Defining Management Domains

<strong>The Four Hs</strong>

Management decisions and activities related to salmon have often been in response to identifiable threats to salmon health and population status. One common framework used to describe impacts to salmon are the “Four Hs”: Habitat, Hydropower, Harvest, and Hatcheries. Many aspects of salmon management are associated with governance and agreements rather than one of the other H’s, which also could be categorized by a fifth H as “Handshaking.” While this categorization structure of salmon concerns presents a good starting point, it is too broad and oversimplified considering the range of legislation involved.

<br>

<strong>International Union for Conservation of Nature (IUCN) Threat Categories</strong>

Human systems can be categorized as direct threats to biodiversity more specifically for conservation strategies (Salafsky et al 2024). The International Union for Conservation of Nature (IUCN) and Conservation Measures Partnership (CMP) has developed a hierarchical grouping of threats used globally by conservation practitioners. IUCN direct threat categories are widely used to assess threats as part of IUCN red list assessments, as well as, in status assessments for Pacific salmon populations conducted by the Committee for the Status of Endangered Wildlife in Canada (COSEWIC). 

For this study, the <a href=" https://docs.google.com/spreadsheets/d/1yfm7ua9hQJpjycx6FYQJ6Jy5LA4bP0XP61EW3-sX8ZY/edit?gid=179026760#gid=179026760" style="color:#668899">IUCN - CMP Direct Threats Classification 4.0</a> was used to categorize legislation by direct threats to salmon conservation. Level 2 threat classification was used as the finest resolution that could be applied legislative clause consistently. 

<br>

<strong>Management Domains</strong>

Legislative clauses that influence Pacific salmon governance in British Columbia are not always associated with conservation threats. As well, some clauses may involve all threats. For example, statutes that outline species status or determine ecological reserves have relevance to how salmon are managed, but can’t be attributed to any one IUCN threat category. Pacific salmon life history also ranges over a broad geography of marine and freshwater habitat (including terrestrial watershed aspects), touching on almost every conservation threat available. “Management Domains” were identified through this study as a clearer and more comprehensive alignment between regulatory administrative concern and conservation threat categories (Figure 1).

```{r flowchart-image, echo=FALSE, results='asis', message=FALSE}
# Load library quietly
suppressPackageStartupMessages(library(here))

# Define image path
image_path <- here("Management Domain Flowchart.png")

# Render image only if it exists
if (file.exists(image_path)) {
  cat(paste0(
    '<div style="text-align: center;">',
    '<img src="', image_path, '" alt="Flowchart" style="max-width: 85%; height: auto;"><br>',
    '<p><em>Figure 1. Flowchart showing alignment of Pacific salmon threat factors (Four H\'s) with Management Domain in relation to legislation and jurisdiction.</em></p>',
    '</div>'
  ))
} else {
  cat(paste0("<p style='color:red;'>❌ Image file not found: ", image_path, "</p>"))
}


```

Management domains represent a reframing of multi-jurisdictional issues by administrative concern rather than specific agency, which can mainstream conservation strategies focused on threats to salmon life history. Management domains can be defined as a “separation of administrative concerns,” and it is important that each domain is categorized distinctly to minimize grey areas and blurring between individual domains or the exercise would fail to be useful. Aside from aspects of governance and organizational structure, 12 specific management domains were delineated for the purpose of this exercise.


```{r management-domain-grid, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)

# Define path to the database
db_path <- here::here("output", "legislation.db")
if (!file.exists(db_path)) stop("Database not found at: ", db_path)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Query unique label_value entries from paragraph_label_table where label_type is 'Management Domain'
query <- "
  SELECT DISTINCT label_value 
  FROM paragraph_label_table 
  WHERE label_type = 'Management Domain' AND label_value IS NOT NULL
  ORDER BY label_value
"
domains <- dbGetQuery(conn, query)$label_value

# Disconnect from the database
dbDisconnect(conn)

# Define CSS for button grid
cat("
<style>
.domain-grid {
  display: grid;
  grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
  gap: 10px;
  margin-top: 20px;
}
.domain-button {
  background-color: #668899;
  color: white;
  border: 1px solid #ccc;
  padding: 10px;
  text-align: center;
  border-radius: 5px;
  font-family: sans-serif;
  cursor: pointer;
  transition: background-color 0.3s;
}
.domain-button:hover {
  background-color: #996666;
}
</style>
")

# Render buttons in grid
cat("<div class='domain-grid'>")
for (domain in domains) {
  cat(sprintf("<div class='domain-button'>%s</div>", domain))
}
cat("</div>")
```

<br>

<strong>Scope</strong>

The process of matching management domains and IUCN threat categories to legislative clauses lacks a measure of relevance to Pacific salmon management. For example, a clause that states catch limits for Chinook salmon would be much more specific to salmon management than a clause that states that a minister shall include ecological integrity objectives in a park management plan. Both clauses influence salmon governance and management, but one clause is more direct. A “Scope” classification was developed as part of this study to parallel management domains and categorize clauses based on their direct application to Pacific salmon. The scope was determined by the specificity of the language used in the section. The scope categories were defined as follows:

<div style="margin-left: 30px; padding-left: 10px">
<ul style='list-style-type: none'>
<li>1 – Salmon: refers to salmon specifically</li>
<li>2 – Fish: refers to fish that include salmon</li>
<li>3 – Habitat: refers to any habitat that includes salmon habitat</li>
<li>4 – Governance: refers to administrative processes and structure</li>
</ul>
</div>

The scope assignment process is linked but occurs independent of the management domain and IUCN threat assignment process. All management domains align with scope, but not the IUCN threats. The “1 – Salmon” scope was assigned based solely on salmon keywords (salmon, Chinook, sockeye, etc.), regardless of management domain or IUCN threat. The “2 – Fish” and “3 – Habitat” scopes align directly to IUCN threats (unless already classified as “1 – Salmon”). The “4 – Governance” scope was applied lastly in a separate matching process.

<br>

### Assigning Management Domains

Assessment of legislative context is prone to subjective interpretation. The inclusion and exclusion of both the overall acts and the clauses within the acts creates points of assumption and bias. For this exercise we cast as wide a net as possible and created a process that is publicly accessible, automated, and repeatable. Other researchers may repeat this process as future legislation and language changes. The process steps are:

<div style="margin-left: 30px; padding-left: 10px">
<ol>
<li>Legislation accumulation and review (CanLii)</li>
<li>Parse legislation by subclause, section, and heading (R)</li>
<li>Match management domain and IUCN threat keywords to subclauses (R)</li>
<li>Match clause type keywords to subclauses (R)</li>
<li>Assign Scope (R)</li>
</ol>
</div>

<br>

<strong>1) Legislation Accumulation and Review</strong>

We focused our review on established Canadian federal and British Columbian provincial legislation (acts and regulations) for inclusion in the management domain process. We did not include specific departmental policy frameworks and processes throughout this exercise unless specifically referenced in the statutes because these executive strategies are prone to changes in the political, administrative, and budgetary landscape. Indigenous legislation, Yukon Territorial legislation, municipal bylaws, and international conventions were not included explicitly. Although this other legislation is pivotal in the management of Pacific salmon, considering this regionally diverse administration was outside the scope of this legislative review in terms of feasibility. However .

The <a href="https://www.canlii.org/" style="color:#668899">Canadian Legal Information Institute website, CanLII</a>, provides an excellent resource for searching federal and provincial acts and regulations by keywords. We created a systematic process for collecting legislation by:

<div style="margin-left: 30px; padding-left: 10px">
<ol>
<li>querying Pacific salmon specific terms (salmon, chinook, sockeye, etc.), which we called "Type A Legislation,"</li>
<li>extracting adjacent keywords from the Type A Legislation (word frequency analysis),</li>
<li>querying new keywords and IUCN Threat keywords in CanLII to find "Type B Legislation" (doesn't mention salmon by name),</li>
<li>filtering legislation regionally (excluding provinces other than BC), and</li>
<li>downloading HTML files for relevant legislation directly from government websites.</li>
</ol>
</div>

Through the CanLII website, the user may look up the keywords queried in each act and regulation chosen. This allows the user to quickly assess the context of the word in the legislation and determine if the section is relevant to Pacific salmon. The full list of legislation reviewed in this process is available in Appendix 5.2.

<br>

<strong>2) Parsing Legislation from HTML Files</strong>

The decentralized overlapping patchwork of Pacific salmon legislation in Canada creates a dynamic such that each act and regulation may contain a variety of administrative responsibilities, possibly attributed to different agencies. However, each act and regulation is comprised of sections and subsections that contain individual clauses that are more specific to a particular management domain. Typological device structure and definitions were derived from the Government of Canada’s <a href="https://www.justice.gc.ca/eng/rp-pr/csj-sjc/legis-redact/legistics/p3p1.html" style="color:#668899">“Legistics Paragraphing”</a> (Department of Justice Canada 2022). For the purpose of this review, we used the individual clause (Figure 2) as the sample unit for assignment into management domains and other attributes.  

```{r legistics-paragraphing, echo=FALSE, results='asis'}
library(here)

# Define image path
image_path <- here("Legistics Paragraphing.png")

# Check if image exists
if (file.exists(image_path)) {
  # Manually create figure with caption below
  cat('<div style="text-align: center;">')
  cat('<img src="', image_path, '" alt="Diagram" style="max-width: 85%; height: auto;"><br>', sep = "")
  cat('<p><em>Figure 2. Fisheries Act (federal) examples of paragraph parsing from HTML code.</em></p>')
  cat('</div>')
} else {
  cat("<p style='color:red;'>❌ Image file not found: ", image_path, "</p>")
}
```

Legislation presented on Canadian and British Columbian government websites--and corresponding HTML code--is structured such that each clause and subclause is separated individually as a paragraph element. HTML code for paragraph elements is generally consistent across legislation, but section, subsection, and heading elements are not. As shown in Figure 2, when we refer to a “clause” we are including subclauses, but also sections, subsections, and opening words (chapeau or umbrella) when they are stand-alone paragraph elements in the HTML. We parsed the HTML files using R and the "rvest" package to extract the text of each clause, then grouped by section and heading.

<br>

<strong>3) Building a Database Compendium</strong>

The parsed legislation was stored in a SQLite database with datatables including:

<div style="margin-left: 30px; padding-left: 10px">
  <ul>
  <li>Legislation Metadata Table
  <ul>
    <li> Jurisdiction</li>
    <li> Act Name</li>
    <li> Legislation Name</li>
    <li> Legislation Type</li>
    </ul></li>
    <li>Paragraph Table
<ul>
<li>Section</li>
<li>Heading</li>
<li>Paragraph</li>
</ul></li>
    <li>Paragraph Label Table
<ul>
<li>Paragraph</li>
<li>Label Type</li>
<li>Label Value</li>
<li>Keyword</li>
<li>Scope</li>
</ul>
</li>
  </ul>
</div>

Storing the data as a database rather than one large spreadsheet allows for a more dynamic and versatile system for querying, while streamlining the parsing and collection process.

<br>

<strong>4) Assigning Management Domains and IUCN Threats</strong>

The Management Domains and IUCN threats were assigned to each section based on a keyword list developed from the IUCN threat categories. The keywords were used to identify specific sections and subsections within the legislation. Each section was then assigned to one or more Management Domain and IUCN threat based on the presence of these keywords.Iteratively, we adjusted the keyword list to ensure that it accurately captured the scope of each Management Domain and IUCN threat. The keyword list is available in Appendix II.

<br>

## Legislative Factors by Threat Categorization

### Clause Types
Similar the Management Domain assignment, we created a Clause Type keyword list to assess the nature or purposes of the clauses. Using language patterns within the legislation, we grouped Clause Type by:

```{r clause-type-list, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(DBI)
library(RSQLite)
library(here)

# Define database path and confirm it exists
db_path <- file.path(here("output"), "legislation.db")
if (!file.exists(db_path)) {
  cat("<p style='color:red;'>❌ Database file not found at: ", db_path, "</p>")
} else {
  # Connect to the database
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # Retrieve unique Clause Type label_values
  query <- "
    SELECT DISTINCT label_value
    FROM paragraph_label_table
    WHERE label_type = 'Clause Type' AND label_value IS NOT NULL
    ORDER BY label_value
  "
  clause_values <- dbGetQuery(conn, query)$label_value
  
  # Disconnect from the database
  dbDisconnect(conn)
  
  # Output as unordered HTML list
  if (length(clause_values) > 0) {
    cat("<div style='margin-top: 20px;'>")
    cat("<h4>Clause Type Categories Identified:</h4>")
    cat("<ul style='padding-left: 20px;'>")
    cat(paste0("<li>", clause_values, "</li>", collapse = "\n"))
    cat("</ul>")
    cat("</div>")
  } else {
    cat("<p>No Clause Type values found in the database.</p>")
  }
}

```

<br>

## Biological and Ecological Concerns in a Current Salmon Management Context

**Description of Wild Salmon Policy and Conservation Units

### Conservation Units and RPAs

**Summary of RPAs and connection of IUCN threats to CUs and DUs

<br>

### NOAA Ecological Concerns and RAMs

*RAMs background and NOAA Ecological Concerns
**Why also use ecological concerns? - life history stages

<br>

# RESULTS

A summary of results from the iterations of the automated compendium are shown below. The complete compendium can also be queried using the "legislation.db" SQLite database, which is available in the "output" folder of this repository. The keyword matching categorization process was applied to 196 acts and regulations chosen as relevant to salmon and habitat management. Broken down to the clause level, the categorization process matched the following totals by scope:

```{r management-domain-tally, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(DBI)
library(RSQLite)
library(here)
library(data.table)

db_path <- file.path(here("output"), "legislation.db")
if (!file.exists(db_path)) {
  cat("<p style='color:red;'>❌ Database file not found at: ", db_path, "</p>")
} else {
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # ✅ Step 1: Get Management Domain-labeled paragraphs directly from label_value
  domain_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS management_domain
    FROM paragraph_label_table
    WHERE label_type = 'Management Domain' AND label_value IS NOT NULL
  "))
  
  # 📄 Step 2: Load paragraph metadata
  paragraph_meta <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, Section, legislation_id
    FROM LegislationParagraphs
    WHERE Section IS NOT NULL
  "))
  
  dbDisconnect(conn)
  
  # 🔗 Step 3: Merge with metadata
  merged <- merge(domain_labels, paragraph_meta, by = "paragraph_id", all.x = TRUE)
  
  # 📊 Step 4: Tally by management_domain
  domain_tally <- merged[, .(
    clauses = uniqueN(paragraph_id),
    sections = uniqueN(paste(legislation_id, Section, sep = "_")),
    statutes = uniqueN(legislation_id)
  ), by = management_domain][order(management_domain)]
  
  # 📦 Totals
  total_clauses <- sum(domain_tally$clauses)
  total_sections <- sum(domain_tally$sections)
  total_statutes <- uniqueN(merged$legislation_id)
  
  # 🖥️ Output
  if (nrow(domain_tally) > 0) {
    cat("<ul style='list-style-type: none; padding-left: 0; margin: 0;'>\n")
    for (i in seq_len(nrow(domain_tally))) {
      cat(sprintf("<li><strong>%s:</strong> %d clauses | %d sections | %d statutes</li>\n",
                  domain_tally$management_domain[i],
                  domain_tally$clauses[i],
                  domain_tally$sections[i],
                  domain_tally$statutes[i]))
    }
    cat(sprintf("<li style='margin-top: 10px;'><strong>Total:</strong> %d clauses | %d sections | %d statutes</li>\n",
                total_clauses, total_sections, total_statutes))
    cat("</ul>\n")
  } else {
    cat("<p>No Management Domain labels found.</p>\n")
  }
}


```

<br>

### Management Domains

Management domains were identified within the 177 of the 196 collected statutes. Figure X below shows a count of clauses by management domain split by jurisdiction. Pollution, Spatial Designation, and Water Use and Watercourse Modifications show the highest number of clauses specific to those categories. Climate Change and Natural Disasters received the fewest matches. 

```{r paragraph-count-by-management-domain, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(ggplot2)

# 📁 Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")
if (!file.exists(db_path)) {
  cat("<p style='color:red;'>❌ Database file not found at: ", db_path, "</p>")
} else {
  # 🔌 Connect to the database
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # 🗂️ Load paragraph labels filtered by Management Domain
  paragraph_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS management_domain
    FROM paragraph_label_table
    WHERE label_type = 'Management Domain' AND label_value IS NOT NULL
  "))
  
  # 🗂️ Load paragraph and legislation metadata
  paragraph_table <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, legislation_id
    FROM LegislationParagraphs
  "))
  
  legislation_metadata <- data.table(dbGetQuery(conn, "
    SELECT legislation_id, jurisdiction
    FROM LegislationMetadata
  "))
  
  # 🔌 Disconnect from the database
  dbDisconnect(conn)
  
  # 🔗 Join labels with paragraph and legislation metadata
  domain_data <- merge(paragraph_labels, paragraph_table, by = "paragraph_id", all.x = TRUE)
  domain_data <- merge(domain_data, legislation_metadata, by = "legislation_id", all.x = TRUE)
  
  # 📊 Summarize paragraph counts by management_domain and jurisdiction
  domain_counts <- domain_data[, .(
    paragraph_Count = uniqueN(paragraph_id)
  ), by = .(management_domain, jurisdiction)][order(-paragraph_Count)]
  
  # 🎨 Create stacked horizontal bar chart
  ggplot(domain_counts, aes(
    x = reorder(management_domain, paragraph_Count),
    y = paragraph_Count,
    fill = jurisdiction
  )) +
    geom_bar(stat = "identity", color = "black", width = 0.8) +
    coord_flip() +
    scale_fill_manual(
      values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
      guide = guide_legend(reverse = TRUE)
    ) +
    labs(
      title = "Paragraph Count by Management Domain and Jurisdiction",
      x = "Management Domain",
      y = "Paragraph Count",
      fill = "Jurisdiction"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12),
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )
}


```

<p style="text-align: center;">
  <em>Figure X. Count of clauses matched by management domains stacked by jurisdiction.</em>
</p>


<br>

IUCN Level 2 categories are more specific to salmon threats, and the categorization process matched to 18 categories. Figure X below shows a count of clauses matched to IUCN Level 2 categories broken down by jurisdiction. Pollution, water management, and fisheries management were the top three, and it should be noted that Spatial Designation as a management domain does not occur within the IUCN Level 2 threat categories. 

```{r paragraph-count-by-iucn, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=12}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(ggplot2)

# 📁 Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")
if (!file.exists(db_path)) {
  cat("<p style='color:red;'>❌ Database file not found at: ", db_path, "</p>")
} else {
  # 🔌 Connect to the database
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # 🗂️ Load paragraph labels filtered by IUCN
  paragraph_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS iucn_l2
    FROM paragraph_label_table
    WHERE label_type = 'IUCN' AND label_value IS NOT NULL
  "))
  
  # 🗂️ Load paragraph and legislation metadata
  paragraph_table <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, legislation_id
    FROM LegislationParagraphs
  "))
  
  legislation_metadata <- data.table(dbGetQuery(conn, "
    SELECT legislation_id, jurisdiction
    FROM LegislationMetadata
  "))
  
  # 🔌 Disconnect from the database
  dbDisconnect(conn)
  
  # 🔗 Join labels with paragraph and legislation metadata
  iucn_data <- merge(paragraph_labels, paragraph_table, by = "paragraph_id", all.x = TRUE)
  iucn_data <- merge(iucn_data, legislation_metadata, by = "legislation_id", all.x = TRUE)
  
  # 📊 Summarize paragraph counts by IUCN threat and jurisdiction
  iucn_counts <- iucn_data[, .(
    paragraph_Count = uniqueN(paragraph_id)
  ), by = .(iucn_l2, jurisdiction)][order(-paragraph_Count)]
  
  # 🎨 Create stacked horizontal bar chart
  ggplot(iucn_counts, aes(
    x = reorder(iucn_l2, paragraph_Count),
    y = paragraph_Count,
    fill = jurisdiction
  )) +
    geom_bar(stat = "identity", color = "black", width = 0.8) +
    coord_flip() +
    scale_fill_manual(
      values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
      guide = guide_legend(reverse = TRUE)
    ) +
    labs(
      title = "Paragraph Count by IUCN Level 2 Threat and Jurisdiction",
      x = "IUCN Threat",
      y = "Paragraph Count",
      fill = "Jurisdiction"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12),
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )
}

```

<p style="text-align: center;"><em>Figure X. Stacked bar chart showing alignment of Pacific salmon threat factors (IUCN Level 2) with jurisdictional distribution of legislative paragraphs.</em></p>

<br>

### Scope

Using the '1 - Salmon' scope for clauses shows the Type A Legislation along with a comparison of the number of paragraphs by Jurisdiction. The results are shown in the bar chart below, which displays the number of paragraphs for each piece of legislation that is specific to salmon. The chart is color-coded by jurisdiction (Federal or Provincial) to highlight the distribution of paragraphs across different legislative frameworks.

```{r unique-paragraph-count-salmon-scope, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=12}
# 📦 Load libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# 🔌 Connect to the database
conn <- dbConnect(SQLite(), dbname = file.path(here("output"), "legislation.db"))

# 🗂️ Load relevant tables
paragraph_labels <- dbGetQuery(conn, "
  SELECT paragraph_id, scope
  FROM paragraph_label_table
  WHERE label_type = 'Salmon Scope' AND scope = '1 - Salmon'
")

paragraph_table <- dbGetQuery(conn, "
  SELECT paragraph_id, legislation_id
  FROM LegislationParagraphs
")

legislation_metadata <- dbGetQuery(conn, "
  SELECT legislation_id, legislation_name, jurisdiction
  FROM LegislationMetadata
")

# 🔌 Disconnect from the database
dbDisconnect(conn)

# 🔗 Join tables according to schema
salmon_data <- paragraph_labels %>%
  inner_join(paragraph_table, by = "paragraph_id") %>%
  inner_join(legislation_metadata, by = "legislation_id")

# 📊 Summarize unique paragraph counts by legislation and jurisdiction
summary_data <- salmon_data %>%
  group_by(legislation_name, jurisdiction) %>%
  summarise(Unique_paragraphs = n_distinct(paragraph_id), .groups = "drop") %>%
  arrange(desc(Unique_paragraphs))

# 🎨 Font scaling for dashboard readability
font_scale <- 1.2

# 📈 Plot or fallback message
if (nrow(summary_data) > 0) {
  ggplot(summary_data, aes(
  x = Unique_paragraphs,
  y = reorder(legislation_name, Unique_paragraphs),
  fill = jurisdiction
)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c(
    "Federal" = "#996666",
    "Provincial" = "#668899",
    "Unknown" = "#999999"
  )) +
  labs(
    title = "Clause Count by Legislation (Scope: 1 - Salmon)",
    x = "Number of Paragraphs",
    y = "Legislation",
    fill = "Jurisdiction"
  ) +
  theme_minimal(base_size = 14 * font_scale) +
theme(
  plot.title = element_text(size = 14 * font_scale, face = "bold", hjust = 0.5),
  plot.title.position = "plot",  # Ensures title spans full plot width
  plot.margin = margin(t = 20, r = 40, b = 20, l = 20),  # Adds extra space on the right
  axis.text.y = element_text(size = 14 * font_scale),
  axis.text.x = element_text(size = 14 * font_scale),
  legend.position = "bottom"
)
} else {
  cat("<p style='color: red;'>⚠️ No paragraphs found for Salmon Scope '1 - Salmon'.</p>")
}

```

<p style="text-align: center;"><em>Figure X. ___________________.</em></p>

<br>

```{r plot-habitat-domain-heatmap-greygrid, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=12}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(ggplot2)

# 🔌 Connect to the database
db_path <- here("output", "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# 🗂️ Load tables
paragraph_labels       <- as.data.table(dbReadTable(conn, "paragraph_label_table"))
legislation_paragraphs <- as.data.table(dbReadTable(conn, "LegislationParagraphs"))
legislation_metadata   <- as.data.table(dbReadTable(conn, "LegislationMetadata"))
dbDisconnect(conn)

# 🎯 Filter for Management Domain labels with scope "3 - Habitat"
habitat_labels <- paragraph_labels[
  label_type == "Management Domain" & scope == "3 - Habitat",
  .(paragraph_id, management_domain = label_value)
]

# 🔗 Extract IUCN Threats from label_value
iucn_labels <- paragraph_labels[
  label_type == "IUCN",
  .(paragraph_id, iucn_l2 = label_value)
]

# 🔗 Merge habitat and IUCN labels
habitat_paragraphs <- merge(habitat_labels, iucn_labels, by = "paragraph_id")

# 🧭 Add jurisdiction context
paragraph_context <- merge(
  legislation_paragraphs[, .(paragraph_id, legislation_id)],
  legislation_metadata[, .(legislation_id, jurisdiction)],
  by = "legislation_id",
  all.x = TRUE
)
habitat_paragraphs <- merge(habitat_paragraphs, paragraph_context, by = "paragraph_id", all.x = TRUE)

# 📊 Aggregate counts
heatmap_data <- habitat_paragraphs[
  !is.na(iucn_l2) & !is.na(management_domain),
  .(paragraph_count = uniqueN(paragraph_id)),
  by = .(management_domain, iucn_l2)
]

# 🎨 Plot heat map with grey grid background
ggplot(heatmap_data, aes(x = iucn_l2, y = management_domain, fill = paragraph_count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#f0f0f0", high = "#996666") +
  labs(
    title = "Clause Density by Management Domain (Scope: 3 - Habitat)",
    x = "IUCN Level 2 Threat Category",
    y = "Management Domain",
    fill = "Clause Count"
  ) +
  theme_minimal(base_size = 14 * font_scale) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.title.position = "plot",
    plot.margin = margin(t = 20, r = 40, b = 20, l = 20),
    panel.grid.major = element_line(color = "grey80", size = 0.3),
    panel.grid.minor = element_line(color = "grey90", size = 0.2),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

<br>

## Clause Types

**Clause type intro

```{r plot-clause-type-heatmap, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=12}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(ggplot2)

# 🔌 Connect to the database
db_path <- here("output", "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# 🗂️ Load tables
paragraph_labels       <- as.data.table(dbReadTable(conn, "paragraph_label_table"))
legislation_paragraphs <- as.data.table(dbReadTable(conn, "LegislationParagraphs"))
legislation_metadata   <- as.data.table(dbReadTable(conn, "LegislationMetadata"))

# 🔌 Disconnect from the database
dbDisconnect(conn)

# 🎯 Extract Clause Type and Management Domain labels
clause_labels <- paragraph_labels[label_type == "Clause Type", .(paragraph_id, clause_type = label_value)]
domain_labels <- paragraph_labels[label_type == "Management Domain", .(paragraph_id, management_domain = label_value)]

# 🔗 Merge clause and domain labels
labeled_paragraphs <- merge(clause_labels, domain_labels, by = "paragraph_id")

# 🔗 Add jurisdiction context
paragraph_context <- merge(
  legislation_paragraphs[, .(paragraph_id, legislation_id)],
  legislation_metadata[, .(legislation_id, jurisdiction)],
  by = "legislation_id",
  all.x = TRUE
)
labeled_paragraphs <- merge(labeled_paragraphs, paragraph_context, by = "paragraph_id", all.x = TRUE)

# 📊 Aggregate counts
heatmap_data <- labeled_paragraphs[
  !is.na(clause_type) & !is.na(management_domain),
  .(paragraph_count = uniqueN(paragraph_id)),
  by = .(clause_type, management_domain)
]

# 🎨 Plot heat map
ggplot(heatmap_data, aes(x = clause_type, y = management_domain, fill = paragraph_count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#f0f0f0", high = "#668899") +
  labs(
    title = "Clause Type vs. Management Domain",
    x = "Clause Type",
    y = "Management Domain",
    fill = "Clause Count"
  ) +
  theme_minimal(base_size = 14 * font_scale) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    plot.title.position = "plot",
    plot.margin = margin(t = 20, r = 40, b = 20, l = 20),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

```
<p style="text-align: center;"><em>Figure X. _______.</em></p>

<br>

## Biological and Ecological Concerns in a Current Salmon Management Context

<br>

# REFERENCES

```{=html}
<style>
.references {
  padding-left: 40px;
  text-indent: -40px;
}
</style>
```

::: references
Arbeider, M., Ritchie, L., Braun, D., Jenewein, B., Rickards, K., Dionne, K., Holt, C., Labelle, M., Nicklin, P., Mozin, P., Grant, P., Parken, C., and Bailey, R. 2020. Interior Fraser Coho Salmon Recovery Potential Assessment. DFO Can. Sci. Advis. Sec. Res. Doc. 2020/025. xi + 211 p.

Armitage, D., de Loë, R., Plummer, R., 2012. Environmental governance and its implications for conservation practice. Conserv. Lett. 5, 245–255.

Beazley, K.F., and Olive, A. 2021. Transforming conservation in Canada: shifting policies and paradigms. FACETS, 6: 1714–1727. <doi:10.1139/> facets-2021-0144.

Becklumb, P. 2013. Federal and provincial jurisdiction to regulate environmental issues. Ottawa.

Bennett NJ, Roth R, Klain SC, Chan K, Christie P, Clark DA, et al. 2017a. Conservation social science: understanding and integrating human dimensions to improve conservation. Biological Conservation, 205: 93–108. DOI: 10.1016/j.biocon.2016.10.006

Campbell ML, and Thomas VG. 2002. Constitutional impacts on conservation—effects of federalism on biodiversity protection. Environment Policy and Law, 32: 223–233.

Collins L, and Sossin L. 2019. Approach to constitutional principles and environmental discretion in Canada. UBC Law Review, 52(1): 293–343. [online]: Available from digitalcommons.osgoode.yorku.ca

Department of Justice Canada. 2022. Legistics Paragraphing. Government of Canada. Accessed at: https://www.justice.gc.ca/eng/rp-pr/csj-sjc/legis-redact/legistics/p3p1.html in August 2025.

DFO. 2020. Recovery Potential Assessment for 11 Designatable Units of Fraser River Chinook Salmon, Oncorhynchus tshawytscha, Part 1: Elements 1 to 11. DFO Can. Sci. Advis. Sec. Sci. Advis. Rep. 2020/023.

DFO. 2022. Recovery Potential Assessment for Fraser River Sockeye Salmon (Oncorhynchus nerka) – Nine Designatable Units – Part 2: Habitat, Threats Assessment, Mitigation, And Allowable Harm. DFO Can. Sci. Advis. Sec. Sci. Advis. Rep. 2022/003.

Dufresne Y, and Ouellet C. 2019. Conceptualization, measurement and effects of positional issues in the Canadian electoral context. Heliyon, 5(4): e01453. <PMID:31025009>. <DOI:10.1016/j.heliyon>. 2019.e01453

Kraus, D., Murphy, S., and Armitage, D. 2021. Ten bridges on the road to recovering Canada’s endangered species. <doi:10.1139/facets>.

Hutchings JA, Côté IM, Dodson JJ, Fleming IA, Jennings S, Mantua NJ, et al. 2012. Is Canada fulfilling its obligations to sustain marine biodiversity? A summary review, conclusions, and recommendations. Environmental Reviews 20: 353–361. DOI: 10.1139/er-2012-0049

Mahony, A., Challenger, W., Robichaud, D., Wright, H., Bussanich, R., Sharma, R., and Enns, J. 2021. Recovery Potential Assessment for the Okanagan Chinook Salmon (Oncorhynchus tshawytscha) (2019). DFO Can. Sci. Advis. Sec. Res. Doc. 2021/009. ix + 106 p.

Marentette, J.R. and Kronlund, A.R. 2020. A Cross-Jurisdictional Review of International Fisheries Policies, Standards and Guidelines: Considerations for a Canadian Science Sector Approach. Can. Tech. Rep. Fish. Aquat. Sci. 3342: xiii + 169 p.

McCune JL, Harrower WL, Avery-Gomm S, Brogan JM, Csergő A-M, Davidson LNK, et al. 2013. Threats to Canadian species at risk: an analysis of finalized recovery strategies. Biological Conservation, 166: 254–265. DOI: 10.1016/j.biocon.2013.07.006

Ray, J.C., Grimm, J., and Olive, A. 2021. The biodiversity crisis in Canada: failures and challenges of federal and sub-national strategic and legal frameworks. FACETS. 6: 1044–1068. <doi:10.1139/facets>.

Salafsky, Nick, Claire Relton, Bruce E. Young, et al. 2025. Classification of Direct Threats to the Conservation of Ecosystems and Species 4.0. Conservation Biology 39 (3): e14434. https://doi.org/10.1111/cobi.14434.

Salafsky, N., Salzer, D., Stattersfield, A.J., Hilton-Taylor, C., Nuegarten, R., Butchart, S.H.M., Collen, B., Cox, N., Master, L.L., O’Connor, S., Wilkie, D., 2008. A standard lexicon for biodiversity conservation: unified classifications of threats and actions. Conserv. Biol. 22, 897–911.

Swerdfager, Trevor, and Andrea Olive. “Laws Matter: A Foundational Approach to Biodiversity Conservation in Canada.” Edited by David Lesbarreres. FACETS 8 (January 1, 2023): 1–13.

Westwood AR, Otto SP, Mooers A, Darimont C, Hodges KE, Johnson C, et al. 2019. Protecting biodiversity in British Columbia: recommendations for developing species at risk legislation. FACETS, 4(1): 136–160. DOI: 10.1139/facets-2018-0042

:::

<hr>

# APPENDICES

<hr>

## Appendix I - Legislation Accumulation and Review

```{r legislation-count-by-type, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)

# 📁 Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")

# 🔍 Check if database file exists
if (!file.exists(db_path)) {
  cat("<p style='color:red;'>❌ Database file not found at:", db_path, "</p>")
} else {
  
  # 🔌 Connect to the database
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # 📋 Check if required table exists
  available_tables <- dbListTables(conn)
  
  if (!"LegislationMetadata" %in% available_tables) {
    cat("<p style='color:red;'>❌ Table 'LegislationMetadata' not found. Available tables: ",
        paste(available_tables, collapse = ", "), "</p>")
    dbDisconnect(conn)
  } else {
    
    # 📊 Query Acts and Regulations from metadata
    metadata <- dbGetQuery(conn, "
      SELECT jurisdiction, legislation_name, legislation_type
      FROM LegislationMetadata
      WHERE legislation_type IN ('Act', 'Regulations')
    ")
    
    # 🔌 Disconnect from the database
    dbDisconnect(conn)
    
    # ⚠️ Handle empty result
    if (nrow(metadata) == 0) {
      cat("<p style='color:orange;'>⚠️ No data found for Acts or Regulations</p>")
    } else {
      
      # 📈 Count Acts by jurisdiction
      acts_count <- metadata %>%
        filter(legislation_type == "Act") %>%
        group_by(jurisdiction) %>%
        summarise(act_count = n_distinct(legislation_name), .groups = "drop") %>%
        arrange(desc(act_count))
      
      # 📈 Count Regulations by jurisdiction
      regs_count <- metadata %>%
        filter(legislation_type == "Regulations") %>%
        group_by(jurisdiction) %>%
        summarise(reg_count = n_distinct(legislation_name), .groups = "drop") %>%
        arrange(desc(reg_count))
      
      # 📝 Display Acts
      cat("<h3>Acts per Jurisdiction</h3>\n<ul>\n")
      if (nrow(acts_count) > 0) {
        for (i in seq_len(nrow(acts_count))) {
          cat(sprintf("<li><strong>%s:</strong> %d Acts</li>\n", 
                      acts_count$jurisdiction[i], 
                      acts_count$act_count[i]))
        }
      } else {
        cat("<li>No Acts found in the database.</li>\n")
      }
      cat("</ul>\n\n")
      
      # 📝 Display Regulations
      cat("<h3>Regulations per Jurisdiction</h3>\n<ul>\n")
      if (nrow(regs_count) > 0) {
        for (i in seq_len(nrow(regs_count))) {
          cat(sprintf("<li><strong>%s:</strong> %d Regulations</li>\n", 
                      regs_count$jurisdiction[i], 
                      regs_count$reg_count[i]))
        }
      } else {
        cat("<li>No Regulations found in the database.</li>\n")
      }
      cat("</ul>\n")
    }
  }
}

```

<hr>

## Appendix II - CoLAPSE Viewer
The following interactive table contains the full compendium of legislation by Management Domain and IUCN Threat category. Search topics by domain and salmon threats to view applicable legislation.

```{r domain-filter-list, echo=FALSE}
library(htmltools)
library(data.table)

# 🗂️ Extract Management Domain values directly from label_value
domain_labels <- paragraph_labels[label_type == "Management Domain", .(paragraph_id, management_domain = label_value)]

# 🧹 Extract and sort unique domains
domains <- sort(unique(na.omit(domain_labels$management_domain)))

# 🎛️ Render filter UI
if (length(domains) == 0) {
  tags$p(style = "color: orange; font-style: italic;", "⚠️ No management domains available to filter.")
} else {
  tags$div(
    style = "margin-bottom: 20px;",
    
    # 🧭 Header and instructions
    tags$strong("Filter by Management Domain:"),
    tags$span(
      "ℹ️ Click a domain to filter the table. Click 'All' to reset.",
      style = "margin-left: 10px; font-size: 0.9em; color: #555;"
    ),
    
    # 🧵 Button container
    tags$div(
      style = "margin-top: 10px; display: flex; flex-wrap: wrap;",
      
      # 🔄 "All" button to clear filter
      tags$button(
        "All",
        onclick = "
          Reactable.setFilter('act-legislation-summary-table', 'DomainFilter', '');
          document.querySelectorAll('[data-domain-button]').forEach(b => b.classList.remove('active-domain'));
          this.classList.add('active-domain');
          return false;
        ",
        `data-domain-button` = NA,
        style = "
          margin-right: 8px;
          margin-bottom: 8px;
          padding: 6px 12px;
          background-color: #996666;
          color: white;
          border: none;
          border-radius: 4px;
          cursor: pointer;
        "
      ),
      
      # 🎯 Domain-specific buttons
      lapply(domains, function(domain) {
        tags$button(
          htmlEscape(domain),
          onclick = sprintf(
            "Reactable.setFilter('act-legislation-summary-table', 'DomainFilter', '%s');
             document.querySelectorAll('[data-domain-button]').forEach(b => b.classList.remove('active-domain'));
             this.classList.add('active-domain');
             return false;",
            htmlEscape(domain)
          ),
          `data-domain-button` = NA,
          style = "
            margin-right: 8px;
            margin-bottom: 8px;
            padding: 6px 12px;
            background-color: #f9f9f9;
            color: #333;
            border: 1px solid #ccc;
            border-radius: 4px;
            cursor: pointer;
          "
        )
      })
    ),
    
    # 🎨 Style for active button
    tags$style(HTML("
      .active-domain {
        background-color: #0074D9 !important;
        color: white !important;
        border-color: #0074D9 !important;
      }
    "))
  )
}
```

```{r act-legislation-summary-table, results='asis', message=FALSE, warning=FALSE, echo=FALSE}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(reactable)
library(htmltools)

# 🗂️ Step 1: Resolve database path safely
db_path <- here::here("output", "legislation.db")
if (!file.exists(db_path)) stop("Database not found at: ", db_path)

# 🔌 Step 2: Connect and load tables
conn <- dbConnect(SQLite(), dbname = db_path)

legislation_paragraphs <- as.data.table(dbReadTable(conn, "LegislationParagraphs"))
legislation_metadata   <- as.data.table(dbReadTable(conn, "LegislationMetadata"))
paragraph_labels       <- as.data.table(dbReadTable(conn, "paragraph_label_table"))
clause_type_keywords   <- as.data.table(dbReadTable(conn, "clause_type_keywords"))

dbDisconnect(conn)

# 🧩 Step 3: Extract and collapse labels
clause_labels <- paragraph_labels[label_type == "Clause Type", .(paragraph_id, keyword)]
clause_labels <- merge(clause_labels, clause_type_keywords[, .(keyword, clause_type)], by = "keyword", all.x = TRUE)
clause_labels <- clause_labels[
  , .(clause_type = unique(clause_type)[1]),  # Select only one unique clause type per paragraph
  by = paragraph_id
]

domain_labels <- paragraph_labels[label_type == "Management Domain", .(paragraph_id, management_domain = label_value)]
domain_labels <- domain_labels[, .(management_domain = paste(unique(management_domain), collapse = ", ")), by = paragraph_id]

iucn_labels <- paragraph_labels[label_type == "IUCN", .(paragraph_id, iucn_l2 = label_value)]
iucn_labels <- iucn_labels[, .(iucn_l2 = paste(unique(iucn_l2), collapse = ", ")), by = paragraph_id]

scope_labels <- paragraph_labels[label_type %in% c("Management Domain", "Salmon Scope"), .(paragraph_id, scope)]
scope_labels <- scope_labels[, .(scope = paste(unique(scope), collapse = ", ")), by = paragraph_id]

# 🔗 Step 4: Merge all labels
label_data <- Reduce(function(x, y) merge(x, y, by = "paragraph_id", all = TRUE, allow.cartesian = TRUE),
                     list(clause_labels, domain_labels, iucn_labels, scope_labels))

# 🔗 Step 5: Merge with paragraph and metadata
section_data <- merge(
  legislation_paragraphs[, .(paragraph_id, legislation_id, Heading, Section, Paragraph)],
  legislation_metadata[, .(legislation_id, legislation_name, act_name, jurisdiction, legislation_type)],
  by = "legislation_id",
  all.x = TRUE
)

section_data <- merge(section_data, label_data, by = "paragraph_id", all.x = TRUE)

# 🧠 Step 6: Create combined columns
section_data[, Act := paste(jurisdiction, act_name, sep = " | ")]
section_data[, Legislation := paste(legislation_type, legislation_name, sep = " | ")]
section_data[, HeadingSection := paste(Heading, paste0("Section ", Section), sep = " | ")]

# 📊 Step 7: Aggregate by Act + Legislation + HeadingSection
section_summary <- section_data[
  !is.na(Section),
  .(
    DomainFilter = paste(unique(na.omit(management_domain)), collapse = ", "),
    Paragraphs = paste(unique(na.omit(Paragraph)), collapse = "<br><br>"),
    ManagementDomain = paste(unique(na.omit(management_domain)), collapse = ", "),
    IUCNThreat = paste(unique(na.omit(iucn_l2)), collapse = ", "),
    ClauseType = paste(unique(na.omit(clause_type)), collapse = ", "),
    Scope = paste(unique(na.omit(scope)), collapse = ", ")
  ),
  by = .(Act, Legislation, HeadingSection)
]

# 🧠 Step 8: Format Paragraphs with metadata block
section_summary[, Paragraphs := paste(
  Paragraphs,
  "<br><br>",
  "<div style='font-size: 0.85em; color: #333; background-color: #f9f9f9; padding: 8px; border-left: 4px solid #0074D9;'>",
  sprintf(
    "<span style='color:#0074D9; font-weight:bold;'>Management Domain:</span> %s; <span style='color:#0074D9; font-weight:bold;'>IUCN Threat:</span> %s; <span style='color:#0074D9; font-weight:bold;'>Clause Type:</span> %s; <span style='color:#0074D9; font-weight:bold;'>Scope:</span> %s",
    htmlEscape(ManagementDomain),
    htmlEscape(IUCNThreat),
    htmlEscape(ClauseType),
    htmlEscape(Scope)
  ),
  "</div>",
  sep = ""
)]

# 🖥️ Step 9: Render reactable inside a responsive container
htmltools::div(
  style = "width: 100%; overflow-x: auto;",
  reactable(
    section_summary[, .(Act, Legislation, HeadingSection, DomainFilter, Paragraphs)],
    elementId = "act-legislation-summary-table",
    groupBy = c("Act", "Legislation"),
    searchable = TRUE,
    pagination = TRUE,
    striped = TRUE,
    bordered = TRUE,
    resizable = TRUE,
    defaultPageSize = 10,
    defaultColDef = colDef(filterable = TRUE),
    columns = list(
      Act = colDef(name = "Act", filterable = TRUE, minWidth = 150),
      Legislation = colDef(name = "Legislation", filterable = TRUE, minWidth = 200),
      HeadingSection = colDef(name = "Heading & Section", filterable = TRUE, minWidth = 150),
      DomainFilter = colDef(name = "Management Domain", filterable = TRUE, show = FALSE),
      Paragraphs = colDef(
        name = "Paragraphs",
        html = TRUE,
        minWidth = 400,
        style = list(fontSize = "0.85em", whiteSpace = "normal")
      )
    )
  )
)


```
             
<hr>

## Appendix III - Keywords: 

```{r keyword-frequency-by-domain-corrected, echo=FALSE, message=FALSE, warning=FALSE}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(reactable)

# 🔌 Connect to database
db_path <- here::here("output", "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# 🗂️ Load paragraph labels
paragraph_labels <- as.data.table(dbReadTable(conn, "paragraph_label_table"))
dbDisconnect(conn)

# 🎯 Step 1: Extract Management Domain keywords
management_labels <- paragraph_labels[label_type == "Management Domain", .(
  management_domain = label_value,
  keyword
)]

keyword_summary <- management_labels[
  !is.na(management_domain) & !is.na(keyword),
  .(Frequency = .N),
  by = .(management_domain, keyword)
][order(-Frequency)]

# 📊 Step 4: Render reactable
reactable(
  keyword_summary,
  groupBy = "management_domain",
  searchable = TRUE,
  pagination = TRUE,
  striped = TRUE,
  bordered = TRUE,
  highlight = TRUE,
  defaultPageSize = 15,
  columns = list(
    management_domain = colDef(name = "Management Domain", minWidth = 150),
    keyword = colDef(name = "Keyword", minWidth = 200),
    Frequency = colDef(name = "Frequency", align = "center")
  )
)


```

<hr>

## Appendix IV - Case Study: 


<hr>

## Appendix V - Case Study:

<hr>

## Appendix VI - External Resources

https://www.canlii.org/

https://laws-lois.justice.gc.ca/eng/acts/

https://www.bclaws.gov.bc.ca/civix/content/complete/statreg/?xsl=/templates/browse.xsl

IUCN Direct Threats Classification 4.0

https://docs.google.com/spreadsheets/d/1yfm7ua9hQJpjycx6FYQJ6Jy5LA4bP0XP61EW3-sX8ZY/edit?gid=179026760#gid=179026760


