---
title: "Compendium of Legislation Applicable to Pacific Salmon and Ecosystems (CoLAPSE)"
author: "Joe Enns, Cory Lagasse"
date: "`r Sys.Date()`"
output:
  scientific::html:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
    fig_caption: true
    code_folding: hide
    link_citations: true
documentclass: article
---
<!-- Typography and Background -->
<style>
body, h1, h2, h3, h4, h5, h6, p, li, td, th {
  font-family: 'Palatino Linotype', 'Book Antiqua', Palatino, serif;
}
</style>

```{r read-legislation-db, message=FALSE, warning=FALSE, echo=FALSE}
# Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)

# Define path to the SQLite database
db_path <- here::here("output", "legislation.db")

# Check if the database file exists
if (!file.exists(db_path)) {
  stop("Database file not found at: ", db_path)
}

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# List of expected tables
expected_tables <- c("LegislationMetadata", "LegislationParagraphs", "provision_label_table")
available_tables <- dbListTables(conn)

# Verify all expected tables are present
missing_tables <- setdiff(expected_tables, available_tables)
if (length(missing_tables) > 0) {
  dbDisconnect(conn)
  stop("Missing tables in database: ", paste(missing_tables, collapse = ", "))
}

# Read tables into R as data.table objects
legislation_metadata   <- as.data.table(dbReadTable(conn, "LegislationMetadata"))
legislation_paragraphs <- as.data.table(dbReadTable(conn, "LegislationParagraphs"))
provision_labels       <- as.data.table(dbReadTable(conn, "provision_label_table"))

# Disconnect from the database
dbDisconnect(conn)
```

<hr>

# INTRODUCTION

## Background

Resource conservation and stewardship in Canada is governed through a decentralized, multi-sector array of constitutional authority administered through federal, provincial, territorial, and municipal governments (Ray et al 2021; Campbell and Thomas 2002). Stemming from the Constitution Act, legislative control over ecological conservation has been a reactionary process of assigning authority to multiple jurisdictions over new environmental issues as they arise, which creates a complex and overlapping distribution of governance (Becklumb 2013). This patchwork of authority is most prominent in Pacific salmon management due to the complicated geographic nature of their anadromous life cycle, spanning a variety of freshwater and marine ecosystems coast-wide and across international boundaries.

While legislation and programs may often operate on a sector-by-sector and individual resource basis, integrated stewardship relies on scale-appropriate planning and decision making (Ray et al 2021). The need for collaborative, transparent decision-making in salmon management is recognized throughout policies, plans, and initiatives, such as the Pacific Salmon Strategy Initiative and the Wild Salmon Policy. Successful conservation and rebuilding of Pacific salmon therefore requires coordination of activities across discrete administrative responsibilities and regulatory frameworks. However, such coordination requires adequate organizational capacity and systems for effective implementation to ensure that regulatory controls and mitigation activities address underlying threats and factors that limit capacity and reduce productivity.

In this report, we present a framework to relate salmon threats with their legislative and regulatory context within British Columbia. We define discrete areas of administrative responsibility for Pacific salmon as “management domains” and conduct a review of legislation to assess the specificity or scope of each clause in relation to Pacific salmon. Our purpose for creating a management domains framework is to:

<div style="margin-left: 30px; padding-left: 10px">
  <ul>
    <li>better understand the legislative and regulatory context that governs Pacific salmon and their ecosystems in British Columbia;</li>
    <li>relate Pacific salmon threat categories to legislation; and</li>
    <li>categorize discrete human activities by administrative concern to support conservation planning across jurisdictions.</li>
  </ul>
</div>

# METHODS

## Defining Management Domains

Management domains represent a reframing of multi-jurisdictional issues by administrative concern rather than specific agency, which can mainstream conservation strategies focused on threats to salmon life history. Since management domains are defined as a “separation of administrative concerns,” it is important that each domain is categorized distinctly to minimize grey areas and blurring between individual domains or the exercise would fail to be useful. Aside from aspects of governance and organizational structure, 13 specific management domains were delineated for the purpose of this exercise. The domains were developed to comprehensively encompass the legislative context, administrative functions, and anthropogenic threats related to Pacific salmon in British Columbia.

```{r management-domain-grid, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)

# Define path to the database
db_path <- here::here("output", "legislation.db")
if (!file.exists(db_path)) stop("Database not found at: ", db_path)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Query unique management domains
query <- "
  SELECT DISTINCT management_domain
  FROM provision_label_table
  WHERE management_domain IS NOT NULL
  ORDER BY management_domain
"
management_domains <- dbGetQuery(conn, query)$management_domain

# Disconnect from the database
dbDisconnect(conn)

# Define CSS for button grid
cat("
<style>
.domain-grid {
  display: grid;
  grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
  gap: 10px;
  margin-top: 20px;
}
.domain-button {
  background-color: #668899;
  color: white; /* ✅ Font color changed to white */
  border: 1px solid #ccc;
  padding: 10px;
  text-align: center;
  border-radius: 5px;
  font-family: sans-serif;
  cursor: pointer;
  transition: background-color 0.3s;
}
.domain-button:hover {
  background-color: #996666;
}
</style>
")

# Render buttons in grid
cat("<div class='domain-grid'>")
for (domain in management_domains) {
  cat(sprintf("<div class='domain-button'>%s</div>", domain))
}
cat("</div>")  # ✅ Properly closed the outer div
```
<br>
We used the International Union for Conservation of Nature (IUCN) threat categories (Salafsky et al 2008) to classify legislation into anthropogenic activities that affect salmon. IUCN threat categories are widely used to assess threats as part of IUCN red list assessments, as well as in status assessments for Pacific salmon populations conducted by the Committee for the Status of Endangered Wildlife in Canada (COSEWIC). We assigned Level 1 and Level 2 threat classification hierarchies from the IUCN categories into management domains to separate anthropogenic activities into different areas of administrative concern.

Historically, much of the management decisions and activities related to salmon have been in response to identifiable threats to salmon health and population status. One common framework used to describe impacts to salmon are the Four H’s: Habitat, Hydropower, Harvest, and Hatcheries. Each management domain, as defined in this report, intentionally aligns with one of the four H’s. However, there are aspects of salmon management associated with governance and agreements rather than one of the other H’s, which we categorized as “Handshaking.” Because the management domains are centered around salmon/habitat impacts, this allows structurally for a clearer alignment with IUCN threat categories.

```{r flowchart-image, echo=FALSE, results='asis'}
library(here)

# Define image path
image_path <- here("Management Domain Flowchart.png")

# Check if image exists
if (file.exists(image_path)) {
  # Manually create figure with caption below
  cat('<div style="text-align: center;">')
  cat('<img src="', image_path, '" alt="Flowchart" style="max-width: 100%; height: auto;"><br>', sep = "")
  cat('<p><em>Figure 1. Flowchart showing alignment of Pacific salmon threat factors (Four H\'s) with Management Domain in relation to legislation and jurisdiction.</em></p>')
  cat('</div>')
} else {
  cat("<p style='color:red;'>❌ Image file not found: ", image_path, "</p>")
}
```

## Assignment Process

Assessment of legislative context is prone to subjective interpretation. The inclusion and exclusion of both the overall acts and the clauses within the acts creates points of assumption and bias. For this exercise we cast as wide a net as possible and created a process that is publicly accessible, automated, and repeatable. Other researchers may repeat this process as future legislation and language changes. The process steps are:

<div style="margin-left: 30px; padding-left: 10px">
<ol>
<li>Legislation accumulation and review (CanLii)</li>
<li>Parse legislation by subclause, section, and heading (R)</li>
<li>Match management domain and IUCN threat keywords to subclauses (R)</li>
<li>Match clause type keywords to subclauses (R)</li>
<li>Assign Scope (R)</li>
</ol>
</div>

### Legislation Accumulation and Review

We focused our review on established Canadian federal and British Columbian provincial legislation (acts and regulations) for inclusion in the management domain process. We did not include specific departmental policy frameworks and processes throughout this exercise unless specifically referenced in the statutes because these executive strategies are prone to changes in the political, administrative, and budgetary landscape. Indigenous legislation, Yukon Territorial legislation, municipal bylaws, and international conventions were not included explicitly. Although this other legislation is pivotal in the management of Pacific salmon, considering this regionally diverse administration was outside the scope of this legislative review in terms of feasibility. However .

The <a href="https://www.canlii.org/" style="color:#668899">Canadian Legal Information Institute website, CanLII</a>, provides an excellent resource for searching federal and provincial acts and regulations by keywords. We created a systematic process for collecting legislation by:

<div style="margin-left: 30px; padding-left: 10px">
<ol>
<li>querying Pacific salmon specific terms (salmon, chinook, sockeye, etc.), which we called "Type A Legislation,"</li>
<li>extracting adjacent keywords from the Type A Legislation (word frequency analysis),</li>
<li>querying new keywords and IUCN Threat keywords in CanLII to find "Type B Legislation" (doesn't mention salmon by name),</li>
<li>filtering legislation regionally (excluding provinces other than BC), and</li>
<li>downloading HTML files for relevant legislation directly from government websites.</li>
</ol>
</div>

Through the CanLII website, the user may look up the keywords queried in each act and regulation chosen. This allows the user to quickly assess the context of the word in the legislation and determine if the section is relevant to Pacific salmon. The full list of legislation reviewed in this process is available in Appendix 5.2.

### Parsing Legislation from HTML Files

The decentralized overlapping patchwork of Pacific salmon legislation in Canada creates a dynamic such that each act and regulation may contain a variety of administrative responsibilities, possibly attributed to different agencies. However, each act and regulation is comprised of sections and subsections that contain individual clauses that are more specific to a particular management domain. For the purpose of the review, we used the individual clause as the sample unit for assignment into management domains and other corresponding attributes.

```{r legistics-paragraphing, echo=FALSE, results='asis'}
library(here)

# Define image path
image_path <- here("Legistics Paragraphing.png")

# Check if image exists
if (file.exists(image_path)) {
  # Manually create figure with caption below
  cat('<div style="text-align: center;">')
  cat('<img src="', image_path, '" alt="Diagram" style="max-width: 100%; height: auto;"><br>', sep = "")
  cat('<p><em>Figure 2. Fisheries Act (federal) examples of paragraph parsing from HTML code.</em></p>')
  cat('</div>')
} else {
  cat("<p style='color:red;'>❌ Image file not found: ", image_path, "</p>")
}
```

We parsed the HTML files using R and the <code>rvest</code> package to extract the text of each subclause, then grouped by section and heading. Standard acro

The text was then cleaned and standardized to remove any extraneous characters or formatting issues. 

Each section was assigned a unique identifier, and metadata such as the act name, jurisdiction, and legislation type (act or regulation) were recorded.

### Assigning Management Domains and IUCN Threats from Keywords

The Management Domains and IUCN threats were assigned to each section based on a keyword list developed from the IUCN threat categories. The keywords were used to identify specific sections and subsections within the legislation. Each section was then assigned to one or more Management Domain and IUCN threat based on the presence of these keywords.Iteratively, we adjusted the keyword list to ensure that it accurately captured the scope of each Management Domain and IUCN threat. The keyword list is available in Appendix II.

### Assigning Scope

We developed a scope classification system to categorize the sections based on their relevance to Pacific salmon. The scope was determined by the specificity of the language used in the section. The Scope categories were defined as follows:
<div style="margin-left: 30px; padding-left: 10px">
<ol>
<li>Salmon - refers to salmon specifically</li>
<li>Fish - refers to fish that include salmon</li>
<li>Habitat - refers to any habitat that includes salmon habitat</li>
<li>Governance - refers to administrative processes and structure</li>
</ol>
</div>

Scope 1 was assigned separately to sections with the specific salmon keywords. Scope 2-3 were assigned during the Management Domain and IUCN Threat assignment.

### Assigning Clause Type from Keywords

Similar the Management Domain assignment, we created a Clause Type keyword list to assess the nature or purposes of the clauses. Using language patterns within the legislation, we grouped Clause Type by:

```{r clause-type-list, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(DBI)
library(RSQLite)
library(here)

# Define database path and confirm it exists
db_path <- file.path(here("output"), "legislation.db")
if (!file.exists(db_path)) stop("❌ Database file not found at: ", db_path)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Retrieve unique clause_type values
query <- "
  SELECT DISTINCT clause_type
  FROM provision_label_table
  WHERE clause_type IS NOT NULL
  ORDER BY clause_type
"
clause_types <- dbGetQuery(conn, query)$clause_type

# Disconnect from the database
dbDisconnect(conn)

# Output as unordered HTML list
if (length(clause_types) > 0) {
  cat("<ul>")
  cat(paste0("<li>", clause_types, "</li>", collapse = "\n"))
  cat("</ul>")
} else {
  cat("<p>No clause types found in the database.</p>")
}

```

# RESULTS

## Compendium of Legislation Applicable to Pacific Salmon and Ecosystems (CoLAPSE)

A summary of results from the initial iterations of the automated compendium are shown below. As the keywords become less specific to salmon, the number of Provisions increases. The compendium is available as an R object, which can be used to query specific sections and subsections by Management Domain, IUCN Threat, Scope, and Clause Type. The full compendium is available in the <code>Full_legislation_compendium.rds</code> file and can be loaded into R using the <code>readRDS()</code> function. The compendium is also available as a CSV file for download from GitHub.

```{r scope-provision-tally, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(DBI)
library(RSQLite)
library(here)
library(dplyr)

# Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")
if (!file.exists(db_path)) stop("❌ Database file not found at: ", db_path)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Query scope and provision_id from the provision_label_table
query <- "
  SELECT scope, provision_id
  FROM provision_label_table
  WHERE scope IS NOT NULL
"
scope_data <- dbGetQuery(conn, query)

# Disconnect from the database
dbDisconnect(conn)

# Tally unique provision_id counts by scope
scope_count <- scope_data %>%
  group_by(scope) %>%
  summarise(count = n_distinct(provision_id), .groups = "drop") %>%
  arrange(scope)

# Print formatted output
cat("<h3>Provision Count by Scope</h3>")
if (nrow(scope_count) > 0) {
  cat("<ul style='list-style-type: none; padding: 0; margin: 0;'>")
  for (i in seq_len(nrow(scope_count))) {
    cat(paste0("<li><strong>", scope_count$scope[i], ":</strong> ", scope_count$count[i], " Provisions</li>\n"))
  }
  cat("</ul>")
} else {
  cat("<p>No scope provisions found in the database.</p>")
}

```
Using the '1 - Salmon' scope for Provisions shows the Type A Legislation along with a comparison of the number of Provisions by Jurisdiction. The results are shown in the bar chart below, which displays the number of Provisions for each piece of legislation that is specific to salmon. The chart is color-coded by jurisdiction (Federal or Provincial) to highlight the distribution of Provisions across different legislative frameworks.

```{r unique-provision-count-salmon-scope, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = file.path(here("output"), "legislation.db"))

# Load tables
provision_labels <- dbGetQuery(conn, "
  SELECT provision_id, scope 
  FROM provision_label_table 
  WHERE scope = '1 - Salmon'
")
provision_table <- dbGetQuery(conn, "
  SELECT provision_id, legislation_id 
  FROM LegislationParagraphs
")
legislation_metadata <- dbGetQuery(conn, "
  SELECT legislation_id, legislation_name, jurisdiction 
  FROM LegislationMetadata
")

# Disconnect
dbDisconnect(conn)

# Join and summarize
salmon_data <- provision_labels %>%
  inner_join(provision_table, by = "provision_id") %>%
  inner_join(legislation_metadata, by = "legislation_id")

summary_data <- salmon_data %>%
  group_by(legislation_name, jurisdiction) %>%
  summarise(Unique_Provisions = n_distinct(provision_id), .groups = "drop") %>%
  arrange(desc(Unique_Provisions))

# Font scaling for dashboard readability
font_scale <- 1.2

# Plot or fallback message
if (nrow(summary_data) > 0) {
  ggplot(summary_data, aes(x = Unique_Provisions, y = reorder(legislation_name, Unique_Provisions), fill = jurisdiction)) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = c("Federal" = "#996666", "Provincial" = "#668899")) +
    labs(
      title = "Provision Count by Legislation (Scope: 1 - Salmon)",
      x = "Number of Provisions",
      y = "Legislation",
      fill = "Jurisdiction"
    ) +
    theme_minimal(base_size = 14 * font_scale) +
    theme(
      axis.text.y = element_text(size = 14 * font_scale),
      axis.text.x = element_text(size = 14 * font_scale),
      plot.title = element_text(size = 14 * font_scale, face = "bold")
    )
} else {
  cat("<p style='color: red;'>⚠️ No provisions found for Scope '1 - Salmon'.</p>")
}
```

## Summary of Management Domains
Including all scopes, the number of Provisions by Management Domain is shown in the bar chart below. The chart displays the number of Provisions for each Management Domain, which represents the administrative areas of concern for Pacific salmon management. The chart is color-coded by jurisdiction (Federal or Provincial) to highlight the distribution of Provisions across different legislative frameworks.

```{r provision-count-by-management-domain, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Load tables
provision_labels <- dbGetQuery(conn, "
  SELECT provision_id, management_domain
  FROM provision_label_table
  WHERE management_domain IS NOT NULL
")

provision_table <- dbGetQuery(conn, "
  SELECT provision_id, legislation_id
  FROM LegislationParagraphs
")

legislation_metadata <- dbGetQuery(conn, "
  SELECT legislation_id, jurisdiction
  FROM LegislationMetadata
")

# Disconnect from the database
dbDisconnect(conn)

# Join tables properly
domain_data <- provision_labels %>%
  inner_join(provision_table, by = "provision_id") %>%
  inner_join(legislation_metadata, by = "legislation_id")

# Summarize unique provision counts by management_domain and jurisdiction
domain_counts <- domain_data %>%
  group_by(management_domain, jurisdiction) %>%
  summarise(Provision_Count = n_distinct(provision_id), .groups = "drop") %>%
  arrange(desc(Provision_Count))

# Generate stacked horizontal bar chart
ggplot(domain_counts, aes(x = reorder(management_domain, Provision_Count), y = Provision_Count, fill = jurisdiction)) +
  geom_bar(stat = "identity", color = "black", width = 0.9) +
  coord_flip() +
  scale_fill_manual(values = c("Federal" = "#996666", "Provincial" = "#668899")) +
  labs(
    title = "Provision Count by Management Domain",
    x = "Management Domain",
    y = "Provision Count",
    fill = "Jurisdiction"
  ) +
  theme_minimal(base_size = 14 * font_scale) +
    theme(
      axis.text.y = element_text(size = 14 * font_scale),
      axis.text.x = element_text(size = 14 * font_scale),
      plot.title = element_text(size = 14 * font_scale, face = "bold")
    )
```

## Summary of IUCN Threats

```{r label-count-by-iucn-l2, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Query relevant columns from provision_label_table and legislation_metadata
iucn_data <- dbGetQuery(conn, "
  SELECT p.label_id, p.iucn_l2, m.jurisdiction
  FROM provision_label_table p
  JOIN LegislationMetadata m ON p.provision_id = m.legislation_id
  WHERE p.iucn_l2 IS NOT NULL
")

# Disconnect from the database
dbDisconnect(conn)

# Summarize label counts by iucn_l2 and jurisdiction
iucn_counts <- iucn_data %>%
  group_by(iucn_l2, jurisdiction) %>%
  summarise(Label_Count = n(), .groups = "drop") %>%
  arrange(desc(Label_Count))

# Generate stacked horizontal bar chart
ggplot(iucn_counts, aes(x = reorder(iucn_l2, Label_Count), y = Label_Count, fill = jurisdiction)) +
  geom_bar(stat = "identity", color = "black", width = 0.9) +
  coord_flip() +
  scale_fill_manual(values = c("Federal" = "#996666", "Provincial" = "#668899")) +
  labs(
    title = "Label Count by IUCN Threat (Split by Jurisdiction)",
    x = "IUCN Level 2 Threat",
    y = "Label Count",
    fill = "Jurisdiction"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12))
```

## Summary of Clause Types

```{r label-count-by-clause-type, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=8}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")

# Connect to the database
conn <- dbConnect(SQLite(), dbname = db_path)

# Query relevant columns from provision_label_table and legislation_metadata
clause_data_raw <- dbGetQuery(conn, "
  SELECT p.label_id, p.clause_type, m.legislation_name
  FROM provision_label_table p
  JOIN LegislationMetadata m ON p.provision_id = m.legislation_id
  WHERE p.clause_type IS NOT NULL
")

# Disconnect from the database
dbDisconnect(conn)

# Summarize label counts and distinct legislation counts by clause_type
clause_data <- clause_data_raw %>%
  group_by(clause_type) %>%
  summarise(
    Label_Count = n(),
    Legislation_Count = n_distinct(legislation_name),
    .groups = "drop"
  ) %>%
  arrange(desc(Label_Count))

# Generate dot plot
ggplot(clause_data, aes(x = clause_type, y = Label_Count)) +
  geom_point(aes(size = Legislation_Count), color = "#889966", alpha = 0.7) +
  scale_size_continuous(range = c(3, 10)) +
  labs(
    title = "Label Count by Clause Type (Dot Size = Number of Legislation Names)",
    x = "Clause Type",
    y = "Label Count",
    size = "Legislation Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

<hr>

# REFERENCES

```{=html}
<style>
.references {
  padding-left: 40px;
  text-indent: -40px;
}
</style>
```

::: references
Armitage, D., de Loë, R., Plummer, R., 2012. Environmental governance and its implications for conservation practice. Conserv. Lett. 5, 245–255.

Beazley, K.F., and Olive, A. 2021. Transforming conservation in Canada: shifting policies and paradigms. FACETS, 6: 1714–1727. <doi:10.1139/> facets-2021-0144.

Becklumb, P. 2013. Federal and provincial jurisdiction to regulate environmental issues. Ottawa.

Bennett NJ, Roth R, Klain SC, Chan K, Christie P, Clark DA, et al. 2017a. Conservation social science: understanding and integrating human dimensions to improve conservation. Biological Conservation, 205: 93–108. DOI: 10.1016/j.biocon.2016.10.006

Campbell ML, and Thomas VG. 2002. Constitutional impacts on conservation—effects of federalism on biodiversity protection. Environment Policy and Law, 32: 223–233.

Collins L, and Sossin L. 2019. Approach to constitutional principles and environmental discretion in Canada. UBC Law Review, 52(1): 293–343. [online]: Available from digitalcommons.osgoode.yorku.ca

Dufresne Y, and Ouellet C. 2019. Conceptualization, measurement and effects of positional issues in the Canadian electoral context. Heliyon, 5(4): e01453. <PMID:31025009>. <DOI:10.1016/j.heliyon>. 2019.e01453

Kraus, D., Murphy, S., and Armitage, D. 2021. Ten bridges on the road to recovering Canada’s endangered species. <doi:10.1139/facets>.

Hutchings JA, Côté IM, Dodson JJ, Fleming IA, Jennings S, Mantua NJ, et al. 2012. Is Canada fulfilling its obligations to sustain marine biodiversity? A summary review, conclusions, and recommendations. Environmental Reviews 20: 353–361. DOI: 10.1139/er-2012-0049

Marentette, J.R. and Kronlund, A.R. 2020. A Cross-Jurisdictional Review of International Fisheries Policies, Standards and Guidelines: Considerations for a Canadian Science Sector Approach. Can. Tech. Rep. Fish. Aquat. Sci. 3342: xiii + 169 p.

McCune JL, Harrower WL, Avery-Gomm S, Brogan JM, Csergő A-M, Davidson LNK, et al. 2013. Threats to Canadian species at risk: an analysis of finalized recovery strategies. Biological Conservation, 166: 254–265. DOI: 10.1016/j.biocon.2013.07.006

Ray, J.C., Grimm, J., and Olive, A. 2021. The biodiversity crisis in Canada: failures and challenges of federal and sub-national strategic and legal frameworks. FACETS. 6: 1044–1068. <doi:10.1139/facets>.

Salafsky, N., Salzer, D., Stattersfield, A.J., Hilton-Taylor, C., Nuegarten, R., Butchart, S.H.M., Collen, B., Cox, N., Master, L.L., O’Connor, S., Wilkie, D., 2008. A standard lexicon for biodiversity conservation: unified classifications of threats and actions. Conserv. Biol. 22, 897–911.

Swerdfager, Trevor, and Andrea Olive. “Laws Matter: A Foundational Approach to Biodiversity Conservation in Canada.” Edited by David Lesbarreres. FACETS 8 (January 1, 2023): 1–13.

Westwood AR, Otto SP, Mooers A, Darimont C, Hodges KE, Johnson C, et al. 2019. Protecting biodiversity in British Columbia: recommendations for developing species at risk legislation. FACETS, 4(1): 136–160. DOI: 10.1139/facets-2018-0042

https://www.justice.gc.ca/eng/rp-pr/csj-sjc/legis-redact/legistics/p3p1.html
:::

<hr>

# APPENDICES

<hr>

## Legislation Accumulation and Review

```{r legislation-count-by-type, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)

# Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")

# Check if database file exists
if (!file.exists(db_path)) {
  cat("<p style='color:red;'>❌ Database file not found at:", db_path, "</p>")
} else {
  
  # Connect to the database
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # Check if table exists
  available_tables <- dbListTables(conn)
  
  if (!"LegislationMetadata" %in% available_tables) {
    cat("<p style='color:red;'>❌ Table 'LegislationMetadata' not found. Available tables:", 
        paste(available_tables, collapse = ", "), "</p>")
    dbDisconnect(conn)
  } else {
    
    # Query relevant columns from LegislationMetadata
    metadata <- dbGetQuery(conn, "
      SELECT jurisdiction, legislation_name, legislation_type
      FROM LegislationMetadata
      WHERE legislation_type IN ('Act', 'Regulations')
    ")
    
    # Disconnect from the database
    dbDisconnect(conn)
    
    # Check if data was retrieved
    if (nrow(metadata) == 0) {
      cat("<p style='color:orange;'>⚠️ No data found for Acts or Regulations</p>")
    } else {
      
      # Count Acts by Jurisdiction
      acts_count <- metadata %>%
        filter(legislation_type == "Act") %>%
        group_by(jurisdiction) %>%
        summarise(count = n_distinct(legislation_name), .groups = "drop") %>%
        arrange(desc(count))
      
      # Count Regulations by Jurisdiction
      regs_count <- metadata %>%
        filter(legislation_type == "Regulations") %>%
        group_by(jurisdiction) %>%
        summarise(count = n_distinct(legislation_name), .groups = "drop") %>%
        arrange(desc(count))
      
      # Print formatted output for Acts
      if (nrow(acts_count) > 0) {
        cat("<h3>Acts per Jurisdiction</h3>\n<ul>\n")
        for (i in seq_len(nrow(acts_count))) {
          cat(sprintf("<li><strong>%s:</strong> %d Acts</li>\n", 
              acts_count$jurisdiction[i], 
              acts_count$count[i]))
        }
        cat("</ul>\n\n")
      } else {
        cat("<h3>Acts per Jurisdiction</h3>\n<p>No Acts found in the database.</p>\n\n")
      }
      
      # Print formatted output for Regulations
      if (nrow(regs_count) > 0) {
        cat("<h3>Regulations per Jurisdiction</h3>\n<ul>\n")
        for (i in seq_len(nrow(regs_count))) {
          cat(sprintf("<li><strong>%s:</strong> %d Regulations</li>\n", 
              regs_count$jurisdiction[i], 
              regs_count$count[i]))
        }
        cat("</ul>\n")
      } else {
        cat("<h3>Regulations per Jurisdiction</h3>\n<p>No Regulations found in the database.</p>\n")
      }
    }
  }
}
```

<hr>

## CoLAPSE Viewer
The following interactive table contains the full compendium of legislation by Management Domain and IUCN Threat category. Search topics by domain and salmon threats to view applicable legislation.

```{r domain-filter-list, echo=FALSE}
library(htmltools)

# Get unique management domains
domains <- sort(unique(na.omit(provision_labels$management_domain)))

# Create clickable filter list with a "Clear" option
tags$div(
  style = "margin-bottom: 20px;",
  tags$strong("Filter by Management Domain:"),
  tags$ul(
    style = "list-style-type: none; padding-left: 0;",
    # Add "All" option to clear filter
    tags$li(
      style = "display: inline-block; margin-right: 10px;",
      tags$a(
        href = "#",
        "All",
        onclick = "Reactable.setFilter('act-legislation-summary-table', 'DomainFilter', ''); return false;",
        style = "font-weight: bold; color: #0074D9;"
      )
    ),
    # Add domain-specific filters
    lapply(domains, function(domain) {
      tags$li(
        style = "display: inline-block; margin-right: 10px;",
        tags$a(
          href = "#",
          htmlEscape(domain),
          onclick = sprintf(
            "Reactable.setFilter('act-legislation-summary-table', 'DomainFilter', '%s'); return false;",
            htmlEscape(domain)
          ),
          style = "color: #333;"
        )
      )
    })
  )
)
```

```{r act-legislation-summary-table, results='asis', message=FALSE, warning=FALSE, echo=FALSE}
# Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(reactable)
library(htmltools)

# Step 1: Resolve database path safely
db_path <- here::here("output", "legislation.db")
if (!file.exists(db_path)) stop("Database not found at: ", db_path)

# Step 2: Connect and load tables
conn <- dbConnect(SQLite(), dbname = db_path)

legislation_paragraphs <- as.data.table(dbReadTable(conn, "LegislationParagraphs"))
legislation_metadata   <- as.data.table(dbReadTable(conn, "LegislationMetadata"))
provision_labels       <- as.data.table(dbReadTable(conn, "provision_label_table"))

dbDisconnect(conn)

# Step 3: Merge tables
section_data <- merge(
  legislation_paragraphs[, .(provision_id, legislation_id, Heading, Section, Paragraph)],
  legislation_metadata[, .(legislation_id, legislation_name, act_name, jurisdiction, legislation_type)],
  by = "legislation_id",
  all.x = TRUE
)

section_data <- merge(
  section_data,
  provision_labels[, .(provision_id, management_domain, iucn_l2, clause_type, scope)],
  by = "provision_id",
  all.x = TRUE
)

# Step 4: Create combined columns
section_data[, Act := paste(jurisdiction, act_name, sep = " | ")]
section_data[, Legislation := paste(legislation_type, legislation_name, sep = " | ")]
section_data[, HeadingSection := paste(Heading, paste0("Section ", Section), sep = " | ")]

# Step 5: Aggregate by Act + Legislation + HeadingSection
section_summary <- section_data[
  !is.na(Section),
  .(
    DomainFilter = paste(unique(na.omit(management_domain)), collapse = ", "),
    Paragraphs = paste(
      paste(na.omit(Paragraph), collapse = "<br><br>"),
      "<br><br>",
      "<div style='font-size: 0.9em; color: #333; background-color: #f9f9f9; padding: 8px; border-left: 4px solid #0074D9;'>",
      sprintf(
        "<strong>management_domain:</strong> <span style='color: #0074D9;'>%s</span><br>
         <strong>iucn_l2:</strong> <span style='color: #0074D9;'>%s</span><br>
         <strong>clause_type:</strong> <span style='color: #0074D9;'>%s</span><br>
         <strong>scope:</strong> <span style='color: #0074D9;'>%s</span>",
        htmlEscape(paste(unique(na.omit(management_domain)), collapse = ", ")),
        htmlEscape(paste(unique(na.omit(iucn_l2)), collapse = ", ")),
        htmlEscape(paste(unique(na.omit(clause_type)), collapse = ", ")),
        htmlEscape(paste(unique(na.omit(scope)), collapse = ", "))
      ),
      "</div>",
      sep = ""
    )
  ),
  by = .(Act, Legislation, HeadingSection)
]

# Step 6: Render reactable (fully static)
reactable(
  section_summary,
  elementId = "act-legislation-summary-table",
  groupBy = c("Act", "Legislation"),
  searchable = TRUE,
  pagination = TRUE,
  striped = TRUE,
  bordered = TRUE,
  resizable = TRUE,
  defaultPageSize = 10,
  defaultColDef = colDef(filterable = TRUE),
  columns = list(
    Act = colDef(name = "Act", filterable = TRUE, minWidth = 200),
    Legislation = colDef(name = "Legislation", filterable = TRUE, minWidth = 300),
    HeadingSection = colDef(name = "Heading & Section", filterable = TRUE, minWidth = 200),
    DomainFilter = colDef(name = "Management Domain", filterable = TRUE, minWidth = 150),
    Paragraphs = colDef(name = "Paragraphs", html = TRUE, minWidth = 500)
  )
)

```
             
<hr>

## Example: The Fisheries Act

<hr>

## Example: The Water Sustainability Act
