---
title: "An interface between statutes and salmon: the Legislation Applicable to Pacific Salmon and Ecosystems (LAPSE) framework"
author: "Joe Enns, Cory Lagasse, Daniel Gillis, Marina Wright"
date: "`r Sys.Date()`"
output:
  bookdown::word_document2:
    reference_docx: "template.docx"
    number_sections: false
    fig_caption: true
    toc: true
    toc_depth: 3
language:
  label:
    fig: "Figure. "
    tab: "Table."
bibliography: LAPSE_primary_publication_references.bib
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)
```

```{r setup-journal-figures, include=FALSE}
## Load Libraries for Journal Figures ----
library(here)
library(DBI)
library(RSQLite)
library(data.table)
library(ggplot2)
library(stringr)
library(scales)

## Connect to Database and Load Tables ----
db_path <- here("output", "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

legislation_table <- as.data.table(dbReadTable(conn, "LegislationMetadata"))
paragraph_table <- as.data.table(dbReadTable(conn, "LegislationParagraphs"))
label_table <- as.data.table(dbReadTable(conn, "paragraph_label_table"))

# Load agencies table for Sankey charts (if available)
agencies_table <- tryCatch({
  as.data.table(dbReadTable(conn, "agencies"))
}, error = function(e) NULL)

dbDisconnect(conn)
```

# Abstract

Pacific salmon (*Oncorhynchus* spp.) populations are generally declining in Canada, while the number of populations monitored continues to drop, despite considerable attention towards policies and recovery programs for salmon conservation. Wild-origin salmon abundance declines are recognized as a consequence of numerous aspects of human activity, including the loss or degradation of habitat, climate change, hatchery pressures, and harvest. Addressing these threats requires coordinated action across a fragmented governance landscape where federal jurisdiction over fisheries intersects with provincial authority over freshwater habitat, creating regulatory complexity that impedes effective stewardship. Practitioners often lack accessible frameworks for navigating the legislative architecture governing salmon and their ecosystems. To address this gap, we present the Legislation Applicable to Pacific Salmon and Ecosystems (LAPSE) framework, which is a systematic process and open-source tool that bridges regulatory language with standardized threat classifications. We accumulated 196 Canadian federal and British Columbian provincial statutes relevant to Pacific salmon stewardship and parsed them by individual clause using automated HTML extraction. Using keyword-matching algorithms, we assigned clauses by "management domains" that we define as discrete areas of administrative concern. These management domains align with the *International Union for Conservation of Nature–Conservation Measures Partnership* (IUCN-CMP) *Direct Threats Classification version 4.0*. Additionally, we categorized clauses by salmon-specific scope (from direct salmon references to general governance) and clause type (e.g., prohibition, authorization, designation). Our analysis reveals that spatial designation and pollution-related provisions are most prevalent across both jurisdictions. Fisheries (harvest) management of salmon is primarily federal, while water use regulations are mostly provincial, which reflects a structural disconnect between harvest control and habitat protection. Case studies linking Fisheries and Oceans Canada's (DFO) Recovery Potential Assessment threat rankings for assessed Designatable Units to corresponding legislation demonstrate the framework's utility for identifying regulatory pathways relevant to specific threats. The LAPSE framework provides salmon stewardship practitioners with navigability of the legal landscape, enabling more informed evaluation of regulatory tools available to address threats and supporting the development of coordinated, inter-jurisdictional recovery strategies.

# Introduction

## The patchwork of Pacific salmon governance in Canada

Eurocentric colonial stewardship of natural resources in Canada is administered through a decentralized, multi-sector array of federal, provincial, and territorial constitutional authority [@ray2021; @campbell2002]. Stemming from the *Constitution Act*, legislative control over fisheries and ecological stewardship has been a reactionary process of assigning authority through two orders of government (federal and provincial) for new environmental issues as they arise, which creates a complex and overlapping governance distribution [@becklumb2013]. Provinces have jurisdiction over most wildlife within their borders; however, the federal government has jurisdiction over wildlife on federal land (e.g. national parks), as well as, migratory birds and aquatic species including marine mammals and fish [@becklumb2013; @kraus2021]. The dual roles – arising from the adaptation of British Common Law to the Canadian situation – can lead to confusion around management prerogatives [@thompson1974] when jurisdictional authority is rarely the responsibility of one level of government [@campbell2002].

The patchwork of legislative authority is most prominent in Pacific salmon management due to the various populations' geographically widespread, anadromous life cycles that spans international boundaries and freshwater and marine ecosystems. While salmon populations and salmon fisheries are regulated federally, water use and watercourse alterations and the corresponding aquatic habitat is managed by the provinces. However, watercourse management is not strictly provincial jurisdiction. The federal government regulates navigable waters in oceans, lakes, rivers, and streams even within provincial boundaries [@becklumb2013; @swerdfager2023]. Federal fisheries governance also includes the quality of fish-bearing waters, and both federal and provincial governments have authority to designate protected areas.

Acts and regulations in British Columbia might be designed to address one broad category of human activity, but this activity can cover many aspects of salmon stewardship. For example, the provincial *Forest and Range Practices Act* [@frpa2002] is specific to forestry practices and planning, along with cattle ranging, but in terms of salmon threats, these activities impact many aspects including stream hydrology, water temperature, effluent, riparian habitat destruction, watercourse changes, [@ulaski2025], and the spread of invasive species.

Projects that influence environmental factors may be subject to both provincial and federal environmental assessments during the planning stages [@becklumb2013]. Natural resource legislation and programs often operate on a sector-by-sector and individual resource basis where integrated stewardship relies on scale-appropriate planning and decision making [@ray2021]. Provinces generally decide when, or if, to regulate on matters of environmental concern [@campbell2002]. British Columbia is even more decentralized than other provinces, especially in terms of species at risk legislation, because British Columbia’s relevant provisions are spread over the greatest number of statutes and regulations [@bogetti2025].

## Complexity in addressing Pacific Salmon population declines

Despite considerable attention towards policies and recovery programs for salmon conservation and governance (e.g. *Canada’s Policy for Conservation of Wild Pacific Salmon and* the *Pacific Salmon Strategy Initiative;* @dfo2005), Pacific salmon populations are generally declining across Canada, while the number of populations monitored continues to drop [@price2017; @atkinson2024]. The causes of Pacific salmon population declines in Canada are varied and have been the subject of numerous major studies and reviews over the last few decades (e.g. the *Commission of Inquiry into the Decline of Sockeye Salmon in the Fraser River (Canada)* [@cohen2012]; *Report of the Standing Committee on Fisheries and Oceans* [@mcdonald2021]; and the *State of the Canadian Pacific salmon: response to changing climate and habitats*[@grant2019]).

The wild-origin salmon abundance declines are recognized as a consequence of numerous aspects of human activity, including the loss or degradation of habitat, climate change, hatchery pressures, and harvest [@bisson2009; @grant2019; @cohen2012; and @mcdonald2021]. In response, commercial fisheries have been curtailed in recent decades [@delvalle2024] as Fisheries and Oceans Canada (DFO) aims to manage fisheries to preserve biodiversity in a manner consistent with the *Wild Salmon Policy* and the "precautionary approach" [@dfo2005]. Commercial salmon fisheries are one aspect that is directly managed by DFO as legislated under the federal *Fisheries Act* [@fisheriesact1985], and subsequent regulations, which prescribe how, when, and where fisheries harvest may occur within Canadian waters with less of the jurisdictional ambiguity. Fisheries also result in a quantifiable mortality of salmon, therefore, impacts and targets can be more accurately measured and managed (assuming adequate monitoring information is collected), and can adjust to changing abundances. This ability to cut fisheries harvest in-season, while the other aspects that impact salmon population productivity and habitat capacity remain and continue to compound, creates a situation where harvest cutbacks (and the corresponding negative economic impacts) have little effect on recovering the salmon populations.

-   **Cory - do we need more citations for end of this paragraph? Do you have some?**

Threats to salmon habitat require a much more coordinated, inter-jurisdictional and inter-departmental approach to evaluate and address. Freshwater habitat, including water quantity (discharge and timing) and water quality (temperature, chemistry, and pollution), are impacted by actions and conditions throughout the entire watershed. Impacts from activities are often indirect, requiring monitoring and long-term studies to identify with confidence [@ulaski2025]. Local activities, such as forestry, agriculture, and urbanization, act cumulatively and compound with climate changes to degrade salmon watersheds [@moore2025] such that the regulation of threats to salmon freshwater habitat in British Columbia requires the application of multiple acts and regulations across both federal and provincial jurisdiction, where the impacts from activities and the benefits of mitigation are uncertain and spread out over time and space. For example, terms specific to pollution (effluent, wastewater, deleterious substances, etc.) occur in 18 Canadian federal acts and 22 British Columbian provincial acts, where the implementation of these acts are the responsibility of eight Canadian federal ministries and nine British Columbian provincial ministries [@departmentofjusticecanada2026; @provinceofbritishcolumbia2025].

-   **Marina - I revised the paragraph to hopefully summarize what you'd suggested in your comment. My original paragraph was vague and unfinished. We don't really mention specific gov't branches and structures since they can change arbitrarily. Hopefully, this still covers what you were recommending.**

Assessment frameworks have been developed amidst this uncertainty and complexity to evaluate anthropogenic activities and limiting factors for the survival and recovery of Pacific salmon populations [@chalifour2022; @salafsky2008; @dfo2021]. These frameworks are varied in their approaches, but aim to be comprehensive in assessing the wide range of concerns that may impact salmon across their life cycle to inform status assessments, recovery planning, and conservation strategies. One widely used risk assessment framework for Pacific salmon is the International Union for Conservation of Nature (IUCN) and Conservation Measures Partnership (CMP) Direct Threats Classification [@salafsky2008], which is applied for Recovery Potential Assessments as part of *Species at Risk Act* listing processes and for Committee on the Status of Endangered Wildlife in Canada (COSEWIC**)** status assessments [@arbeider2020; @dionne2023; @doutaz2023].

Rebuilding plans have been developed under the *Fisheries Act* Fish Stock Provisions [@dfo2022] for a small number of Pacific salmon populations including Okanagan Chinook [@DFO2022] and West Coast Vancouver Island Chinook [@dfo2025]. Risk assessments are intended to guide the development of rebuilding plans and mitigation actions with the ultimate aim of reversing declines and improving conservation status, assuming there is agreement upon the most important threats. A key challenge is identifying actions that are impactful and feasible across the wide geographic range and diverse life history of Pacific salmon. In many cases, dispersed authority, regulatory complexities, ambiguous responsibilities, ill-defined standards of management, and unclear conservation objectives across jurisdictions become barriers to effective and timely action.

## Why study salmon legislation: purpose and objectives

The need for collaborative, coordinated action to address Pacific salmon declines is recognized by governments throughout policies, plans, and initiatives including the *Wild Salmon Policy* [@dfo2005] and the *Trilateral declaration on the urgent need to address wild Pacific salmon populations in British Columbia* [@dfo2024]. Coordinated action requires adequate organizational capacity and systems for implementation to ensure that regulatory controls and mitigation activities address underlying threats to salmon and their habitat. However, linking threats and recovery actions to the relevant regulatory systems can be complex because salmon are often affected by multiple threats that each span across agencies and governmental jurisdiction. In this study, we seek to decode the complexity between threats to salmon and corresponding environmental legislation by creating an interface between regulatory language and anthropogenic threat categories.

Laws formally codify societal values and expectations [@swerdfager2023] by governing human activity. Salmon population management administered through legislative authority, therefore, relies on a tertiary system where laws control human activity, which impacts the quantity and quality of salmon habitat, which affects the capacity of the habitat to produce salmon. Because this system of control pivots around human activity, we also focused this study and tools around human activity, employing the hierarchical categorization structure developed by the IUCN-CMP. However, not all aspects of salmon management are specific to threats from human activity. Topics such as restoration, species status, spatial designation (parks), and overall governance administration may not connect to a direct human threat, or may encompass all the threats. In order to capture all these threat and non-threat related categories in our legislation-salmonscape interface, we created a schema of "management domains" that we define as "discrete areas of administrative concern."

We developed and outlined the *Legislation Applicable to Pacific Salmon and their Ecosystems* (LAPSE) as a process and tool to parse Canadian federal and British Columbian provincial acts and regulations relevant to Pacific salmon, then use keyword phrase relationships to match clauses (paragraphs) to our management domains and corresponding IUCN-CMP threat categories. By creating an extensive, open-source, and accessible database linking legislation to salmon, we aim to facilitate a better understanding of the legislative and regulatory context through which salmon and their ecosystems are governed in British Columbia, while providing a framework to support systematic and critical reviews of strengths and shortcomings of regulations with respect to salmon threats.

# Management domains and the International Union for Conservation of Nature- Conservation Measures Partnership threat categories

@salafsky2008 recognized that using a common language and a standard lexicon is essential for conservation science. Threats are a fundamental conservation concept, and can be defined as "current or future human activities or altered natural phenomena that affect one or more ecosystems or species of interest" [@salafsky2024]. Legislated recovery planning also relies on a description of identified threats [@mccune2013]. The IUCN and the CMP began developing classifications of direct threats in the early 2000s, which were then compiled and developed further through an iterative process resulting in the *IUCN – CMP Direct Threats Classification Version 4.0* [@salafsky2024] used for this exercise. The IUCN Direct Threats Classification uses a hierarchical, exclusive, comprehensive, and expandable structure of three levels similar to the Linnean system of grouping by family, genera, and species [@salafsky2008]. First-level categories (Level 1) are the broadest, and are subdivided into second-level categories (Level 2), which are further subdivided into the most specific, third-level categories (Level 3).

Canadian and British Columbian legislation has been enacted on a sector-by-sector basis, which doesn't exactly align well with the IUCN-CMP Level 1 categories. The IUCN-CMP Level 2 categories fit well with corresponding management domains and legislative categories; however, could be grouped for simplicity and to be more intuitive (Figure \@ref(fig:fig-flowchart-image)). IUCN-CMP Level 3 categories were not used in this assessment.

```{r fig-flowchart-image, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Flowchart showing alignment of IUCN-CMP Level 2 threat categories with Management Domain in relation to legislation and jurisdiction."}
library(here)
library(knitr)

# Define image path
image_path <- here("assets", "Management Domain Flowchart.png")

# Render image only if it exists
if (file.exists(image_path)) {
  include_graphics(image_path)
} else {
  plot.new()
  text(0.5, 0.5, paste("❌ Image file not found:", image_path))
}
```

Lumping and splitting between IUCN-CMP Levels 1 and 2 categories was required to match threats with legislation (Table \@ref(tab:management-domain-threat-table)), which is why the management domain categories were required. For example, aquaculture and hatcheries are important aspects that impact Pacific salmon populations. This topic spans two IUCN-CMP Level 1 categories: 2. Agriculture & Aquaculture and 8. Invasive / Other Problematic Species, Genes & Pathogens, but would be more specifically aligned with the IUCN-CMP Level 2 categories: 2.4 Marine & Freshwater Aquaculture and 8.3 Introduced Genetic Material. For clarity and better alignment, we created a management domain called Aquaculture and Hatcheries. A similar misalignment occurs when assessing the provincial *Forest And Range Practices Act*. Since the management domain categories were derived as close as possible to the IUCN-CMP threat categories (a mix of Level 1 and Level 2), most can be viewed as equivalent.

```{r management-domain-threat-table}
library(flextable)

# Query tables from the database
conn <- dbConnect(SQLite(), dbname = db_path)
threat_table <- as.data.table(dbReadTable(conn, "management_domain_threat_table"))
label_table <- as.data.table(dbReadTable(conn, "paragraph_label_table"))
dbDisconnect(conn)

# Get unique IUCN L2 values that matched paragraphs
matched_iucn <- unique(label_table[label_type == "IUCN" & !is.na(label_value), .(iucn_l2 = label_value)])

# Filter threat table: include rows with matched IUCN values OR rows with NA IUCN (governance domains)
# Check for both string "NA" and actual NA values
threat_table_filtered <- threat_table[
  iucn_l2 %in% matched_iucn$iucn_l2 | iucn_l2 == "NA" | is.na(iucn_l2)
]

# Create flag for sorting: NA IUCN values should come last
threat_table_filtered[, is_na_iucn := iucn_l1 == "NA" | is.na(iucn_l1)]

# Replace NA values (both string and actual) with "NA" string for display
threat_table_filtered[iucn_l1 == "NA" | is.na(iucn_l1), iucn_l1 := "NA"]
threat_table_filtered[iucn_l2 == "NA" | is.na(iucn_l2), iucn_l2 := "NA"]

# Order: non-NA IUCN first (by management domain, then IUCN), then NA IUCN rows
setorder(threat_table_filtered, is_na_iucn, management_domain, iucn_l2, iucn_l1)

# Select columns for display (Management Domain, IUCN L2, IUCN L1)
display_table <- threat_table_filtered[, .(
  `Management Domain` = management_domain,
  `IUCN-CMP Level 2` = iucn_l2,
  `IUCN-CMP Level 1` = iucn_l1
)]

# Create flextable with caption and row grouping
ft <- flextable(display_table) |>
  set_caption("Management domains and corresponding IUCN - CMP threat categories.") |>
  merge_v(j = "Management Domain") |>
  merge_v(j = "IUCN-CMP Level 1") |>
  valign(valign = "top") |>
  width(j = 1, width = 1.8) |>
  width(j = 2, width = 2.5) |>
  width(j = 3, width = 2.3) |>
  theme_booktabs() |>
  bold(part = "header") |>
  hline(part = "body", border = officer::fp_border(color = "gray80"))

ft
```

# Parsing legislation: a process for clause categorization

Assessment of legislative context is prone to subjective interpretation. The inclusion and exclusion of both the overall acts and the clauses within the acts creates points of assumption and bias. For this exercise we cast as wide a net as possible and created a process that is publicly accessible, automated, and repeatable if legislation changes. The process steps are: 1) accumulate and review legislation (CanLII); 2) parse legislation by clause (paragraph), section, and heading; 3) match management domain and IUCN-CMP threat keywords to clauses; 4) assign salmon-specific scope categories; 5) match clause type keywords to clauses.

## Legislation accumulation and review

Administrative concerns specific to natural resource stewardship are also often split between ministries at the federal and provincial level, which can further decentralize responsibility. Figures \@ref(fig:fig-sankey-federal) and \@ref(fig:fig-sankey-provincial) show a compilation of federal and provincial acts relevant to Pacific salmon stewardship with assigned ministries. Relevant acts were compiled through a step-wise keyword search process using the Canadian Legal Information Institute (CanLII) website [@canadianlegalinformationinstitute2025]. The ministries were referenced from the federal *Table of Public Statutes and Responsible Ministers* [@departmentofjusticecanada2026] and the British Columbian provincial *Acts - Minsters' Responsibilities* [@provinceofbritishcolumbia2025]*.* The ministry and minister names shown may not necessarily represent the current cabinet names, which can be changed arbitrarily by the government of the day. The table represents the minister and ministry names as they are stated in the legislation and corresponding minister responsibility tables, which are less prone to short-term changes.

```{r fig-sankey-federal, echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=12, dpi=300, fig.cap = "List of Canadian federal acts relevant to Pacific salmon and habitat governance with assigned ministries."}

library(ggalluvial)
library(stringr)
library(ggrepel)

if (!is.null(agencies_table)) {
  # Compute unique section count per legislation_id
  section_counts <- paragraph_table[
    !is.na(Section),
    .(section_count = uniqueN(Section)),
    by = legislation_id
  ]
  
  # Merge with legislation metadata
  section_counts <- merge(
    section_counts,
    legislation_table[, .(legislation_id, act_name, legislation_name, jurisdiction)],
    by = "legislation_id",
    all.x = TRUE
  )
  
  # Aggregate by act_name to avoid duplicate joins
  section_counts <- section_counts[
    , .(section_count = sum(section_count)), by = .(act_name, jurisdiction)
  ]
  
  # Merge with agencies
  sankey_data <- merge(
    agencies_table,
    section_counts,
    by = "act_name",
    all.x = TRUE
  )
  
  # Filter to Federal and aggregate
  sankey_federal <- sankey_data[!is.na(section_count) & jurisdiction == "Federal"]
  sankey_federal <- sankey_federal[, .(section_count = sum(section_count)), by = .(act_name, agency)]
  
  # Proportionally attribute sections across agencies
  sankey_federal[, n_agencies := .N, by = act_name]
  sankey_federal[, section_count := section_count / n_agencies]
  
  # Set all counts to 1 to show equal-sized strata (connections only)
  sankey_federal[, section_count := 1]
  
  if (nrow(sankey_federal) > 0) {
    # Order acts and agencies alphabetically for consistent layout
    act_order <- sankey_federal[, .(total = sum(section_count)), by = act_name][order(act_name)]
    agency_order <- sankey_federal[, .(total = sum(section_count)), by = agency][order(agency)]
    
    sankey_federal[, act_name := factor(act_name, levels = act_order$act_name)]
    sankey_federal[, agency := factor(agency, levels = agency_order$agency)]
    
    # Create alluvial plot - grey bands, colored sides
    p <- ggplot(sankey_federal,
           aes(axis1 = act_name, axis2 = agency, y = section_count)) +
      geom_alluvium(fill = "gray80", width = 1/12, color = "gray60", linewidth = 0.2, alpha = 0.7) +
      geom_stratum(width = 1/12, fill = "#996666", color = "black", linewidth = 0.3) +
      # Legislation labels on the left - with ggrepel for no overlap
      geom_text_repel(
        stat = "stratum",
        aes(label = after_stat(
          ifelse(x == 1, str_wrap(as.character(stratum), width = 30), "")
        )),
        size = 3.2,
        hjust = 1,
        nudge_x = -0.15,
        direction = "y",
        segment.color = "gray60",
        segment.size = 0.2,
        min.segment.length = 0,
        box.padding = 0.2,
        point.padding = 0.1,
        force = 1,
        max.overlaps = Inf,
        lineheight = 0.9,
        family = "Times"
      ) +
      # Agency labels on the right - with ggrepel for no overlap
      geom_text_repel(
        stat = "stratum",
        aes(label = after_stat(
          ifelse(x == 2, str_wrap(as.character(stratum), width = 30), "")
        )),
        size = 3.2,
        hjust = 0,
        nudge_x = 0.15,
        direction = "y",
        segment.color = "gray60",
        segment.size = 0.2,
        min.segment.length = 0,
        box.padding = 0.2,
        point.padding = 0.1,
        force = 1,
        max.overlaps = Inf,
        lineheight = 0.9,
        family = "Times"
      ) +
      scale_x_discrete(limits = c("Legislation", "Ministry"), expand = c(0.35, 0.35)) +
      theme_minimal(base_size = 12, base_family = "Times") +
      theme(
        text = element_text(color = "black", size = 10, family = "Times"),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(t = 10, r = 120, b = 10, l = 120, unit = "pt")
      )
    
    # Explicitly print the plot
    print(p)
  }
}
```

```{r fig-sankey-provincial, echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=12, dpi=300, fig.cap="List of British Columbian provincial acts relevant to Pacific salmon and habitat governance with assigned ministries."}

library(ggalluvial)
library(stringr)

if (!is.null(agencies_table)) {
  # Compute unique section count per legislation_id
  section_counts <- paragraph_table[
    !is.na(Section),
    .(section_count = uniqueN(Section)),
    by = legislation_id
  ]
  
  # Merge with legislation metadata
  section_counts <- merge(
    section_counts,
    legislation_table[, .(legislation_id, act_name, legislation_name, jurisdiction)],
    by = "legislation_id",
    all.x = TRUE
  )
  
  # Aggregate by act_name to avoid duplicate joins
  section_counts <- section_counts[
    , .(section_count = sum(section_count)), by = .(act_name, jurisdiction)
  ]
  
  # Merge with agencies
  sankey_data <- merge(
    agencies_table,
    section_counts,
    by = "act_name",
    all.x = TRUE
  )
  
  # Filter to Provincial and aggregate
  sankey_provincial <- sankey_data[!is.na(section_count) & jurisdiction == "Provincial"]
  sankey_provincial <- sankey_provincial[, .(section_count = sum(section_count)), by = .(act_name, agency)]
  
  # Proportionally attribute sections across agencies
  sankey_provincial[, n_agencies := .N, by = act_name]
  sankey_provincial[, section_count := section_count / n_agencies]
  
  # Set all counts to 1 to show equal-sized strata (connections only)
  sankey_provincial[, section_count := 1]
  
  if (nrow(sankey_provincial) > 0) {
    # Order acts and agencies alphabetically for consistent layout
    act_order <- sankey_provincial[, .(total = sum(section_count)), by = act_name][order(act_name)]
    agency_order <- sankey_provincial[, .(total = sum(section_count)), by = agency][order(agency)]
    
    sankey_provincial[, act_name := factor(act_name, levels = act_order$act_name)]
    sankey_provincial[, agency := factor(agency, levels = agency_order$agency)]
    
    # Create alluvial plot - grey bands, colored sides
    p <- ggplot(sankey_provincial,
           aes(axis1 = act_name, axis2 = agency, y = section_count)) +
      geom_alluvium(fill = "gray80", width = 1/12, color = "gray60", linewidth = 0.2, alpha = 0.7) +
      geom_stratum(width = 1/12, fill = "#668899", color = "black", linewidth = 0.3) +
      # Legislation labels on the left - wrapped
      geom_text(
        stat = "stratum",
        aes(label = after_stat(
          ifelse(x == 1, str_wrap(as.character(stratum), width = 30), "")
        )),
        size = 3.2,
        hjust = 1,
        nudge_x = -0.1,
        lineheight = 0.9,
        family = "Times"
      ) +
      # Agency labels on the right - wrapped
      geom_text(
        stat = "stratum",
        aes(label = after_stat(
          ifelse(x == 2, str_wrap(as.character(stratum), width = 30), "")
        )),
        size = 3.2,
        hjust = 0,
        nudge_x = 0.1,
        lineheight = 0.9,
        family = "Times"
      ) +
      scale_x_discrete(limits = c("Legislation", "Ministry"), expand = c(0.3, 0.3)) +
      theme_minimal(base_size = 12, base_family = "Times") +
      theme(
        text = element_text(color = "black", size = 10, family = "Times"),
        axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(t = 10, r = 100, b = 10, l = 100, unit = "pt")
      )
    
    # Explicitly print the plot
    print(p)
  }
}
```

We accumulated Canadian federal and British Columbian provincial legislation according to their relevance to Pacific salmon and habitat stewardship. For this study, the term legislation (or statutes) refers to acts, regulations, codes, and orders. Policies are formal statements about how the government of the day plans to execute the legislative requirements [@marentette2020]. Legislation is a better indicator of a jurisdiction's long-term commitment to natural resource stewardship [@bogetti2025] than policies because policies are more prone to changes in the political, administrative, and budgetary landscape, while legislation is subject to due process or a parliamentary vote. Specific departmental policy frameworks and processes were not assessed within this exercise. This study also does not verify the implementation or enforcement in practice of the accumulated statutes by the responsible officials. The accumulated legislation is simply shown as it is written.

Indigenous legislation, Yukon Territorial legislation, municipal bylaws, and international conventions were also not included explicitly. Although this other legislation is pivotal in the stewardship of Pacific salmon, this regionally diverse administration was outside the scope of this legislative review in terms of feasibility and timelines. However, some of these other conventions may overlap with federal or provincial legislation in their governance. For example, the provincial *Spheres of Concurrent Jurisdiction — Environment and Wildlife Regulation* under the *Community Charter* references municipal jurisdiction in relation to the environment and wildlife.

The *Canadian Legal Information Institute* website, CanLII [@canadianlegalinformationinstitute2025], provides an excellent resource for searching federal and provincial legislation by keyword search. We used CanLII through a step-wise, systematic process to query salmon-specific terms (salmon, Chinook, sockeye, etc.). Often, popular language and phrasing used by practitioners may not appear in legal language due to the age and generality of the documents. In order to capture salmon-specific language that actually occurs in the legal language, we extracted other adjacent keywords and phrases from the salmon-specific legislation using a word frequency analysis with the "udpipe" R package [@wijffels2025]. A list of extracted adjacent keywords and phrases were combined with the IUCN threat description wordings, which were then queried in CanLII. The user may quickly view each individual keyword within the chosen statute through the CanLII website. This allows the user to assess the context of the word in the legislation and determine if the section might be relevant to Pacific salmon. Legislation was filtered by region (excluding provinces other than BC), and the statutes' HTML files were downloaded directly from government websites. The number of statutes collected for inclusion in this study is shown in Table \@ref(tab:legislation-summary-table).

```{r legislation-summary-table, echo=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(flextable)

# Connect to the database
db_path <- here("output", "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# Load legislation metadata
legislation <- as.data.table(dbReadTable(conn, "LegislationMetadata"))
dbDisconnect(conn)

# Create count summary by jurisdiction and legislation type
count_table <- legislation[, .N, by = .(Jurisdiction = jurisdiction, `Legislation Type` = legislation_type)]
count_table <- dcast(count_table, Jurisdiction ~ `Legislation Type`, value.var = "N", fill = 0)

# Define desired column order
desired_cols <- c("Jurisdiction", "Act", "Regulations", "Order", "Code")

# Keep only columns that exist in the data, in the desired order
existing_cols <- intersect(desired_cols, names(count_table))
count_table <- count_table[, ..existing_cols]

# Add totals column
count_table[, Total := rowSums(.SD), .SDcols = names(count_table)[-1]]

# Add total row
total_row <- data.table(Jurisdiction = "Total")
for (col in names(count_table)[-1]) {
  total_row[[col]] <- sum(count_table[[col]])
}
count_table <- rbind(count_table, total_row)

# Create flextable for Word output
ft <- flextable(count_table) %>%
  bold(part = "header") %>%
  bold(i = nrow(count_table), part = "body") %>%
  hline(i = nrow(count_table) - 1, border = officer::fp_border(width = 1)) %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  autofit() %>%
  set_caption("Count of legislation by jurisdiction and legislation type.")

ft
```

## Legistics

The decentralized overlapping patchwork of Pacific salmon legislation in Canada creates a dynamic such that each act and regulation may contain a variety of administrative responsibilities, possibly attributed to different agencies. However, each act and regulation is comprised of sections and subsections that contain individual clauses that are more specific to a particular management domain topic. Typological device structure and definitions used in this exercise were derived from the Government of Canada’s “Legistics Paragraphing” [@departmentofjusticecanada2022]. For the purpose of this review, we used the individual clause (Figure \@ref(fig:fig-legistics-paragraphing)) as the sample unit for assignment into management domains and other attributes.

```{r fig-legistics-paragraphing, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "<span style='font-style: italic; font-family: Times New Roman; font-size: 10px'>Example of HTML code paragraph parsing structure for Section 29 of the Canadian federal *Fisheries Act</em>.*"}
library(here)
library(knitr)

# Define image path
image_path <- here("assets", "Legistics Paragraphing.png")

# Check if image exists
if (file.exists(image_path)) {
  include_graphics(image_path)
} else {
  plot.new()
  text(0.5, 0.5, paste("❌ Image file not found:", image_path))
}
```

Legislation presented on Canadian and British Columbian government websites, and corresponding HTML code, is structured such that each clause is separated individually as a paragraph element. HTML code for paragraph elements is generally consistent across legislation, but section, subsection, and heading elements are not. As shown in Figure \@ref(fig:fig-legistics-paragraphing), “clause” in this study can also refer to subclauses, but also sections, subsections, and opening words (chapeau or umbrella) when they are stand-alone paragraph elements. HTML files were parsed using R and the “rvest” package [@wickham2025], extracting the text of each clause, then grouping by section and heading.

## The LAPSE process and tool development

The parsed legislation was stored in a SQLite database with datatables including jurisdiction, act name, legislation name, legislation type, paragraph content, section, heading, management domain, IUCN-CMP threat category, scope, and keywords matched. Storing the data as a database rather than one large spreadsheet allowed for a more dynamic and versatile system for querying, while streamlining the parsing and collection process.

## Keyword and keyphrase development

The management domain and IUCN-CMP threat matching process pivoted around an established list of keywords and phrases relevant to Pacific salmon stewardship. Keywords act as the language link between categorization and the legislative clauses. Keywords must be specific enough to match an IUCN-CMP threat, yet general enough to appear in multiple acts and regulations that span many decades of legal language. We derived keyword lists from the *IUCN - CMP Direct Threats Classification 4.0* examples and definitions; word frequency analyses (unigrams, bigrams, and trigrams) of collected legislation; and iterative researcher input and revision.

The keyword and keyphrase development process required an iterative researcher revision step because of the varied context of language. While the words used in the IUCN-CMP threat definitions may be specific to those threats, they may not be relevant to salmon stewardship and may also have multiple meanings leading to an over-assignment of categories. For example, the word “landfill” is listed on the IUCN-CMP threat examples for 1.2 Commercial & Industrial Areas, in the context of the physical footprint of landfill sites. Much of the collected legislation deals with pollution, in which landfills are mentioned as a source of effluent. However, for salmon habitat, the main threat would come from water effluent, rather than the physical impact of landfill sites adjacent to salmon habitat, and pollution would be a more accurate designation. “Landfill” as a keyword adds many clauses that could be considered outside of the scope of this tool, and so “landfill” was removed from the keyword list. Conversely, the word “riparian” is very relevant to salmon habitat stewardship in British Columbia, yet that word doesn’t appear in the IUCN-CMP threat examples or word frequency analyses of other legislation. “Riparian” was added as a keyword solely from the researchers’ discretion.

Researcher subjectivity can influence the process on two fronts: legislation inclusion, and keyword list development. However, using keyword lists allows for the subjectivity to be acknowledged, and the keyword lists to be reported such that the process may be repeatable. The meaning and context behind keywords, especially single words, can cause mis-categorization in unexpected ways. For example, the word “stock” is an important keyword for salmon stewardship (the Fisheries management domain category). However, as a single word, “stock” can also refer to business holdings, freight cars, and cattle, which do not fit the Fisheries management domain. Revision of “stock” as a keyword is important for the accuracy of categorization. Outputs were reviewed iteratively, and keyword revision was a continuous process to find more relevant bigram or trigram alternatives; remove keywords where sections are already matched with other, more applicable keywords; and search for more specific word variations (plural, past tense, etc.)

# Results

## Management domains

While the number of statutes, sections, or clauses is not a measure of the quality or effectiveness of the legislation [@ray2021], the LAPSE process and tool can be used by practitioners to view the existing acts, regulations, orders, and codes when considering actions or proposals. Management domains were identified within all the collected statutes. Spatial Designation and Pollution were the management domains with the highest section count (Figure \@ref(fig:section-count-by-management-domain)), split almost evenly between federal and provincial jurisdictions. Human Disturbance (specifically, 6.3 Other Human Disturbances, as in species research activities) and Climate Change and Natural Disasters had the fewest section count. Fisheries was in the top five for section count, with a majority of sections under federal jurisdiction. Water Use and Watercourse Modifications was in the top three for section count, with a majority of sections under provincial jurisdiction. This count indicates an underlying possibly dysfunctional relationship where salmon harvest is controlled federally while habitat aspects are governed provincially.

```{r section-count-by-management-domain, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, fig.cap="Count of sections matched by management domain, stacked by jurisdiction."}
# Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(ggplot2)
library(stringr)

# Connect to the database
db_path <- file.path(here("output"), "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# Load paragraph labels filtered by Management Domain
paragraph_labels <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, label_value AS management_domain
  FROM paragraph_label_table
  WHERE label_type = 'Management Domain' AND label_value IS NOT NULL
"))

# Load paragraph and legislation metadata including Section
paragraph_table <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, legislation_id, Section
  FROM LegislationParagraphs
"))

legislation_metadata <- data.table(dbGetQuery(conn, "
  SELECT legislation_id, jurisdiction
  FROM LegislationMetadata
"))

# Disconnect from the database
dbDisconnect(conn)

# Join labels with paragraph and legislation metadata
domain_data <- merge(paragraph_labels, paragraph_table, by = "paragraph_id", all.x = TRUE)
domain_data <- merge(domain_data, legislation_metadata, by = "legislation_id", all.x = TRUE)

# Normalize management domain labels
domain_data[, management_domain := str_squish(management_domain)]

# Summarize counts by unique Section-legislation_id pairs
domain_counts <- domain_data[!is.na(Section), .(
  section_count = uniqueN(paste(Section, legislation_id))
), by = .(management_domain, jurisdiction)]

# Order management domains by total section count (largest to smallest)
domain_order <- domain_counts[, .(total = sum(section_count)), by = management_domain][order(-total)]
domain_counts[, management_domain := factor(management_domain, levels = rev(domain_order$management_domain))]

# Create stacked horizontal bar chart
ggplot(domain_counts, aes(
  x = management_domain,
  y = section_count,
  fill = jurisdiction
)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.8) +
  coord_flip() +
  scale_fill_manual(
    values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    x = "Management Domain",
    y = "Section Count",
    fill = "Jurisdiction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    text = element_text(color = "black"),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

## IUCN-CMP Level 2 threats

IUCN-CMP Level 2 categories are more specific, and the categorization process matched to 25 categories (Figure \@ref(fig:fig-section-count-by-iucn)) that parallel the management domain categories. Multiple IUCN-CMP threats can comprise a management domain; however 9.1 Water-Borne & Other Effluent Pollution matched with the highest number of sections followed by 7.2 Dams & Water Management / Use and 5.4 Fishing, Harvesting & Controlling Aquatic Species. This assessment shows the proportional layout of historic legal attention given to topics, where pollution receives a large amount of attention, while topics such as temperature changes receives very little legal attention.

```{r fig-section-count-by-iucn, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8, fig.cap="Section counts by IUCN-CMP threat category, separated by jurisdiction."}
# 📦 Load required libraries
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(ggplot2)
library(stringr)

# 📁 Define path to the SQLite database
db_path <- file.path(here("output"), "legislation.db")

if (!file.exists(db_path)) {
  # Fallback: show a blank plot with error text so knitting doesn't fail
  plot.new()
  text(0.5, 0.5, paste("Database file not found at:", db_path))
} else {
  # Connect to the database

  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # Load paragraph labels filtered by IUCN
  paragraph_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS iucn_l2
    FROM paragraph_label_table
    WHERE label_type = 'IUCN' AND label_value IS NOT NULL
  "))
  
  # Load paragraph and legislation metadata including Section
  paragraph_table <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, legislation_id, Section
    FROM LegislationParagraphs
  "))
  
  legislation_metadata <- data.table(dbGetQuery(conn, "
    SELECT legislation_id, jurisdiction
    FROM LegislationMetadata
  "))
  
  # Disconnect from the database
  dbDisconnect(conn)
  
  # Join labels with paragraph and legislation metadata
  iucn_data <- merge(paragraph_labels, paragraph_table, by = "paragraph_id", all.x = TRUE)
  iucn_data <- merge(iucn_data, legislation_metadata, by = "legislation_id", all.x = TRUE)
  
  # Normalize IUCN labels
  iucn_data[, iucn_l2 := str_squish(iucn_l2)]
  
  # Summarize counts by unique Section-legislation_id pairs
  iucn_counts <- iucn_data[!is.na(Section), .(
    section_Count = uniqueN(paste(Section, legislation_id))
  ), by = .(iucn_l2, jurisdiction)]
  
  # Order IUCN threats by total section count (largest to smallest)
  iucn_order <- iucn_counts[, .(total = sum(section_Count)), by = iucn_l2][order(-total)]
  iucn_counts[, iucn_l2 := factor(iucn_l2, levels = rev(iucn_order$iucn_l2))]
  
  # Create stacked horizontal bar chart
  ggplot(iucn_counts, aes(
    x = iucn_l2,
    y = section_Count,
    fill = jurisdiction
  )) +
    geom_bar(stat = "identity", color = "black", width = 0.8) +
    coord_flip() +
    scale_fill_manual(
      values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
      guide = guide_legend(reverse = TRUE)
    ) +
    labs(
      title = NULL,
      x = "IUCN Level 2 threat",
      y = "Section count",
      fill = "Jurisdiction"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12),
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )
}
```

## Salmon-specific scope

Regulatory language that can be matched to management domain and IUCN-CMP threat categories is not inherently relevant to Pacific salmon. However, clauses fall within a relevance spectrum. For example, a clause that states catch limits for Chinook salmon would be much more specific to salmon stewardship than a clause that states that a minister shall include ecological integrity objectives in a park management plan. Both clauses influence salmon governance and stewardship, but one clause is more direct. A “scope” classification was developed as part of this study to parallel management domains and categorize clauses based on their direct application to Pacific salmon. Scope categories were determined by the specificity of the keywords and defined as: "1 – Salmon" (refers to Pacific salmon specifically); "2 – Fish" (refers to fish that include salmon); "3 – Habitat" (refers to any habitat that includes salmon habitat); and "4 – Governance" (refers to administrative processes and structure).

The scope assignment process was linked, but occurred independent of the management domain and IUCN-CMP threat assignment process. The “1 – Salmon” scope was assigned based solely on salmon keywords (salmon, Chinook, sockeye, etc.), regardless of management domain or IUCN-CMP threat. The “2 – Fish” and “3 – Habitat” scopes align directly to IUCN-CMP threats (unless already classified as “1 – Salmon”). The “4 – Governance” scope was applied last in a separate matching process. Language specific to Pacific salmon management was found in far fewer sections than indirect, less specific language (Table \@ref(tab:scope-tally)). More federal legislation sections matched with salmon harvest categories than provincial.

```{r scope-tally, echo=FALSE, message=FALSE, warning=FALSE}
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(flextable)

# Connect to database
db_path <- file.path(here("output"), "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# Get scope-labeled paragraphs from Management Domain and Salmon Scope
scope_labels <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, scope
  FROM paragraph_label_table
  WHERE label_type IN ('Management Domain', 'Salmon Scope') AND scope IS NOT NULL
"))

# Load paragraph metadata
paragraph_meta <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, Section, legislation_id
  FROM LegislationParagraphs
  WHERE Section IS NOT NULL
"))

dbDisconnect(conn)

# Merge with metadata
merged <- merge(scope_labels, paragraph_meta, by = "paragraph_id", all.x = TRUE)

# Tally by scope
scope_tally <- merged[, .(
  Statutes = uniqueN(legislation_id),
  Sections = uniqueN(paste(legislation_id, Section, sep = "_")),
  Clauses = uniqueN(paragraph_id)
), by = scope][order(scope)]

# Rename column for display
setnames(scope_tally, "scope", "Scope")

# Calculate totals
total_row <- data.table(
  Scope = "Total",
  Statutes = uniqueN(merged$legislation_id),
  Sections = sum(scope_tally$Sections),
  Clauses = sum(scope_tally$Clauses)
)

# Append total row
scope_tally <- rbind(scope_tally, total_row)

# Create flextable for Word output
ft <- flextable(scope_tally) %>%
  bold(part = "header") %>%
  bold(i = nrow(scope_tally), part = "body") %>%
  hline(i = nrow(scope_tally) - 1, border = officer::fp_border(width = 1)) %>%
  align(j = 2:4, align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  autofit() %>%
  set_caption("Count of statutes, sections, and clauses by scope category.")

ft
```

\
Pacific salmon-specific language ("1 - Salmon") also matched more federal statutes (Figure \@ref(fig:section-count-salmon-scope), specifically, the *British Columbia Sport Fishing Regulations, 1996*; *National Parks of Canada Fishing Regulations*; and *Pacific Fishery Regulations 1993*. The federal *Fisheries Act* does not use the word “salmon” or any salmon-specific keywords at all, only the regulations made under the *Fisheries Act*.

```{r section-count-salmon-scope, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, fig.cap="Count of sections by legislation for scope category '1 - Salmon', coloured by jurisdiction."}
# Load libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = file.path(here("output"), "legislation.db"))

# Load relevant tables
paragraph_labels <- dbGetQuery(conn, "
  SELECT paragraph_id, scope
  FROM paragraph_label_table
  WHERE label_type = 'Salmon Scope' AND scope = '1 - Salmon'
")

paragraph_table <- dbGetQuery(conn, "
  SELECT paragraph_id, legislation_id, Section
  FROM LegislationParagraphs
")

legislation_metadata <- dbGetQuery(conn, "
  SELECT legislation_id, legislation_name, jurisdiction
  FROM LegislationMetadata
")

# Disconnect from the database
dbDisconnect(conn)

# Join tables according to schema
salmon_data <- paragraph_labels %>%
  inner_join(paragraph_table, by = "paragraph_id") %>%
  inner_join(legislation_metadata, by = "legislation_id") %>%
  filter(!is.na(Section))

# Summarize unique Section-legislation_id counts by legislation and jurisdiction
summary_data <- salmon_data %>%
  mutate(section_key = paste(Section, legislation_id)) %>%
  group_by(legislation_name, jurisdiction) %>%
  summarise(Unique_sections = n_distinct(section_key), .groups = "drop")

# Reorder legislation by total section count (largest to smallest)
summary_data <- summary_data %>%
  arrange(desc(Unique_sections)) %>%
  mutate(legislation_name = factor(legislation_name, levels = rev(legislation_name)))

# Create horizontal bar chart
ggplot(summary_data, aes(
  x = Unique_sections,
  y = legislation_name,
  fill = jurisdiction
)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.8) +
  scale_fill_manual(
    values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    x = "Section Count",
    y = "Legislation",
    fill = "Jurisdiction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    text = element_text(color = "black"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

## Clause types as legislative factors

The process of parsing legislation by clause and section, then matching categories by keywords also allows for other possible categorizations. Similar to management domains and IUCN-CMP threats, the nature of the clauses was also matched in terms of the “clause type” (Table \@ref(tab:clause-type-with-domain-tally)). Clauses, in general, fit into groupings by their context and the nature of their intent. For this exercise, the categories and keywords were developed solely by the researchers through a subjective process.

```{r clause-type-with-domain-tally, echo=FALSE, message=FALSE, warning=FALSE}
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(flextable)
# Connect to database
db_path <- file.path(here("output"), "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)
# Get Management Domain and Clause Type labels
label_data <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, label_type, label_value
  FROM paragraph_label_table
  WHERE label_type IN ('Management Domain', 'Clause Type') 
    AND label_value IS NOT NULL
"))
# Load paragraph metadata
paragraph_meta <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, Section, legislation_id
  FROM LegislationParagraphs
  WHERE Section IS NOT NULL
"))
dbDisconnect(conn)
# Extract Management Domain and Clause Type labels separately
domain_labels <- label_data[label_type == 'Management Domain', .(paragraph_id, management_domain = label_value)]
clause_labels <- label_data[label_type == 'Clause Type', .(paragraph_id, clause_type = label_value)]
# Merge to get paragraphs with both labels
merged_labels <- merge(domain_labels, clause_labels, by = "paragraph_id", allow.cartesian = TRUE)
# Merge with paragraph metadata
merged <- merge(merged_labels, paragraph_meta, by = "paragraph_id", all.x = TRUE)
# Create unique section identifier
merged[, section_key := paste(legislation_id, Section, sep = "_")]
# Count unique sections by management domain and clause type
crosstab <- merged[, .(Sections = uniqueN(section_key)), by = .(management_domain, clause_type)]
# Pivot to wide format
crosstab_wide <- dcast(crosstab, management_domain ~ clause_type, value.var = "Sections", fill = 0)
# Rename first column for display
setnames(crosstab_wide, "management_domain", "Management Domain")
# Calculate row totals
clause_cols <- setdiff(names(crosstab_wide), "Management Domain")
crosstab_wide[, Total := rowSums(.SD), .SDcols = clause_cols]
# Order by Total descending
crosstab_wide <- crosstab_wide[order(-Total)]
# Calculate column totals
total_row <- crosstab_wide[, lapply(.SD, sum), .SDcols = c(clause_cols, "Total")]
total_row[, `Management Domain` := "Total"]
setcolorder(total_row, names(crosstab_wide))
# Append total row
crosstab_wide <- rbind(crosstab_wide, total_row)
# Abbreviate long column names to save space
names(crosstab_wide) <- gsub("Administration, Application, & Structure", "Admin.", names(crosstab_wide))
names(crosstab_wide) <- gsub("Authorization & Mandate", "Auth.", names(crosstab_wide))
names(crosstab_wide) <- gsub("Designation", "Desig.", names(crosstab_wide))
names(crosstab_wide) <- gsub("Instruction", "Instr.", names(crosstab_wide))
names(crosstab_wide) <- gsub("Interpretation & Purpose", "Interp.", names(crosstab_wide))
names(crosstab_wide) <- gsub("Licence, Permitting, & Exemptions", "Licence", names(crosstab_wide))
names(crosstab_wide) <- gsub("Prohibition, Restriction, or Limitation", "Prohib.", names(crosstab_wide))
# Create flextable for Word output
ft <- flextable(crosstab_wide) %>%
  # Reduce font size to fit
  fontsize(size = 9, part = "all") %>%
  fontsize(size = 8, part = "header") %>%
  # Bold formatting
  bold(part = "header") %>%
  bold(i = nrow(crosstab_wide), part = "body") %>%
  # Add line above total row
  hline(i = nrow(crosstab_wide) - 1, border = officer::fp_border(width = 1)) %>%
  # Alignment
  align(j = 2:ncol(crosstab_wide), align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  # Fit to page width
  set_table_properties(width = 1, layout = "autofit") %>%
  # Set caption (simplified)
  set_caption("Section counts by management domain and clause type.") %>%
  # Add footnote with abbreviations
  add_footer_lines("Abbreviations: Admin. = Administration, Application, & Structure; Auth. = Authorization & Mandate; Desig. = Designation; Instr. = Instruction; Interp. = Interpretation & Purpose; Licence = Licence, Permitting, & Exemptions; Prohib. = Prohibition, Restriction, or Limitation.") %>%
  # Format the footer
  fontsize(size = 8, part = "footer") %>%
  italic(part = "footer")

ft
```

## Co-occurrence of management domains

The LAPSE keyword matching process allows that multiple management domains (and IUCN-CMP threats) can be matched to clauses and sections. For example the provincial *Land-based Finfish Waste Control Regulation* (*Environmental Management Act*) states in Section 2 that: “The owner of a land-based finfish facility is exempt from section 6 (2) and (3) of the Environmental Management Act with respect to the discharge of wastes from the land-based finfish facility if the owner…” Keywords in this clause matched appropriately with both the “Aquaculture and Hatcheries” and “Pollution” management domains. Assigning only one management domain would over-simplify the classification in terms of salmon management as the clause clearly references both aquaculture and pollution. This section could be important for users searching either aquaculture or pollution regulations for a particular watercourse. This dynamic is not a flaw, but rather a way of showing the complexities of legal language as related to salmon stewardship. Domain co-occurrence within sections allows for an interesting study of how management domains overlap (Figure \@ref(fig:co-occurrence-pairs)).

```{r co-occurrence-pairs, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.cap="Management domain co-occurrence associations by section count."}
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(igraph)
library(ggraph)
library(ggplot2)

# Database connection
db_path <- file.path(here("output", "legislation.db"))
if (!file.exists(db_path)) {
  stop("Database file not found at: ", db_path)
}
conn <- dbConnect(SQLite(), dbname = db_path)

# Query management domain labels
domain_labels <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, label_value AS domain
  FROM paragraph_label_table
  WHERE label_type = 'Management Domain' 
    AND label_value IS NOT NULL
"))

# Query paragraph metadata
paragraph_meta <- data.table(dbGetQuery(conn, "
  SELECT paragraph_id, Section, legislation_id
  FROM LegislationParagraphs
  WHERE Section IS NOT NULL
"))

dbDisconnect(conn)

# Identify unique domains per section
merged <- merge(domain_labels, paragraph_meta, by = "paragraph_id", all.x = TRUE)
section_domains <- merged[, .(domains = list(unique(domain))), 
                          by = .(Section, legislation_id)]

# Generate canonical domain pairs
domain_pairs <- section_domains[sapply(domains, length) >= 2, {
  pairs <- combn(sort(unique(unlist(domains))), 2, simplify = FALSE)
  data.table(
    domain1 = sapply(pairs, `[`, 1),
    domain2 = sapply(pairs, `[`, 2)
  )
}, by = .(Section, legislation_id)]

# Count co-occurrences and filter
pair_counts <- domain_pairs[, .N, by = .(domain1, domain2)]
edges <- pair_counts[N > 50]  # Increased threshold to reduce clutter

# Only create plot if we have data
if (nrow(edges) > 0) {
  # Calculate node sizes based on unique SECTIONS per domain
  all_domains <- unique(c(edges$domain1, edges$domain2))
  node_freq <- merged[domain %in% all_domains, 
                      .(freq = uniqueN(paste(Section, legislation_id))), 
                      by = domain]
  
  # Create graph object
  graph <- graph_from_data_frame(
    d = edges[, .(from = domain1, to = domain2, weight = N)],
    directed = FALSE,
    vertices = node_freq
  )
  
  # Create network plot
  ggraph(graph, layout = "fr") +
    geom_edge_link(
      aes(width = weight, alpha = weight),
      color = "#76948C"
    ) +
    geom_node_point(
      aes(size = freq),
      color = "#4A6A7A",
      alpha = 0.8
    ) +
    geom_node_text(
      aes(label = name),
      size = 3.5,
      repel = TRUE,
      box.padding = 0.5,
      segment.color = "grey70"
    ) +
    scale_edge_width(
      name = "Co-occurrence\n(Sections)",
      range = c(0.3, 2.5)
    ) +
    scale_edge_alpha(
      name = "Co-occurrence\n(Sections)",
      range = c(0.15, 0.95)  # Wider range: weak edges very transparent, strong edges opaque
    ) +
    scale_size(
      name = "Sections per\nDomain",
      range = c(4, 15)
    ) +
    theme_void() +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 9)
    ) +
    guides(
      edge_width = guide_legend(order = 1),
      edge_alpha = guide_legend(order = 1),
      size = guide_legend(order = 2)
    )
} else {
  # Create empty plot with message if no data
  ggplot() +
    annotate("text", x = 0, y = 0, 
             label = "No domain pairs with >50 co-occurrences found",
             size = 5) +
    theme_void()
}
```

# Recovery potential assessments (RPA) and threat ranking case studies

The *Committee on the Status of Endangered Wildlife in Canada* (COSEWIC) assesses the status of wildlife species at risk in Canada, which is the first step in the process of listing a wildlife species under the *Species at Risk Act* (SARA). COSEWIC uses Designatable Units (DUs) to assess populations below the level of a recognized taxonomic species if the populations have attributes that make them both “discrete” and “evolutionarily significant” [@cosewic2023]. COSEWIC outlines DUs within Pacific salmon species according to these criteria for the purposes of its status assessments of these species. When a DU under DFO’s responsibility is designated as threatened, endangered, or extirpated, a DFO-led Recovery Potential Assessment (RPA) process is initiated, which often includes a threat assessment using the same IUCN-CMP threat classification used in this legislative assessment.

DFO's *Wild Salmon Policy* (WSP) outlines salmon management delineated by Conservation Units (CU), which are defined as: “a group of wild salmon sufficiently isolated from other groups that, if extirpated is very unlikely to recolonize naturally within an acceptable time frame, such as a human lifetime or a specified number of salmon generations” [@dfo2005]. In most cases, DUs and CUs are equivalent with some lumping and splitting for specific populations. Any differences are explained within the RPA documents [@dionne2023; @dfo2020; @mahony2019; @arbeider2020; @doutaz2023; @dfo2019]. Since the RPA population and habitat descriptions, along with corresponding threats, are conducted in the context of DUs, we incorporated DUs for the purpose of this study. We only included salmon populations with published RPAs (Table \@ref(tab:du-species-du-name-table)), and only RPA processes that used comparable IUCN-CMP threat categories. For example, Sakinaw Lake sockeye had an RPA published in 2019, but used a "limiting factor" analysis [@ramshaw2019], which did not align with this analysis.

```{r du-species-du-name-table, message=FALSE, warning=FALSE, echo=FALSE}
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(flextable)

# Connect to the database
db_path <- file.path(here("output"), "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# Query the du_list table
du_data <- as.data.table(dbReadTable(conn, "du_list"))

# Disconnect
dbDisconnect(conn)

# Order by species, then by DU number (ascending)
du_data <- du_data[order(species, as.numeric(du_number))]

# Select columns for display
du_display <- du_data[, .(species, du_number, du_name, rpa_citation)]

# Rename columns for display
setnames(du_display, c("species", "du_number", "du_name", "rpa_citation"), c("Species", "DU Number", "Designatable Unit", "RPA Reference"))

# Create flextable for Word output
ft <- flextable(du_display) %>%
  bold(part = "header") %>%
  align(j = c(1, 3, 4), align = "left", part = "all") %>%
  align(j = 2, align = "center", part = "all") %>%
  set_table_properties(width = 1, layout = "autofit") %>%
  set_caption("Designatable Units included in the DFO Recovery Potential Assessment analyses.")
ft
```

<br>

The RPA process identified two IUCN-CMP threat categories as being an "Extreme" or "Extreme-High" ranking (Figure \@ref(fig:du-ranking-heatmap)). 10.1 Geological Events received an extreme ranking, and 7.2 Dams & Water Management / Use received an extreme-high ranking. These two categories were chosen to assess the categories as case studies.

```{r du-ranking-heatmap, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, fig.cap="Threat ranking categories assigned by IUCN-CMP threat categories for Designatable Units of Pacific salmon assessed through the Recovery Potential Assessment process."}
# Load libraries
library(here)
library(data.table)
library(RSQLite)
library(ggplot2)

# Connect to SQLite database
db_path <- file.path(here("output"), "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# Read cu_ranking table
cu_ranking <- as.data.table(dbReadTable(conn, "cu_ranking"))

# Disconnect
dbDisconnect(conn)

# Filter out NA and empty values
cu_ranking <- cu_ranking[
  !is.na(rpa_threat_ranking) & rpa_threat_ranking != "" &
  !is.na(iucn_l2) & iucn_l2 != "" &
  !is.na(du_name) & du_name != "" &
  !is.na(ranking_id)
]

# Convert ranking_id to numeric (if not already)
cu_ranking[, ranking_id := as.numeric(ranking_id)]

# Order rpa_threat_ranking by numeric ranking_id
ordered_levels <- cu_ranking[order(ranking_id), unique(rpa_threat_ranking)]
cu_ranking[, rpa_threat_ranking := factor(rpa_threat_ranking, levels = ordered_levels)]

# Extract leading numbers from iucn_l2 for natural sorting
iucn_levels <- unique(cu_ranking$iucn_l2)
extract_sort_key <- function(x) {
  match <- regmatches(x, regexpr("^[0-9]+(\\.[0-9]+)?", x))
  if (length(match) == 0 || match == "") return(c(Inf, Inf))
  parts <- as.numeric(unlist(strsplit(match, "\\.")))
  if (length(parts) == 1) return(c(parts[1], 0))
  return(c(parts[1], parts[2]))
}
sort_keys <- t(sapply(iucn_levels, extract_sort_key))
iucn_order <- iucn_levels[order(sort_keys[,1], sort_keys[,2])]
cu_ranking[, iucn_l2 := factor(iucn_l2, levels = iucn_order)]

# Aggregate unique du_name count by combination
heatmap_data <- cu_ranking[, .(unique_du_count = uniqueN(du_name)), by = .(rpa_threat_ranking, iucn_l2)]

# Define anchor colors for threat rankings
anchor_colors <- c(
  "Negligible" = "#FFFFCC",
  "Low" = "#FAF49A",
  "Medium" = "#FFD129",
  "High" = "#F75532",
  "Extreme" = "#971C2A"
)

# Create interpolated color palette for all levels
n_levels <- length(ordered_levels)
color_ramp <- colorRampPalette(anchor_colors)(n_levels)
names(color_ramp) <- ordered_levels

# Plot heatmap with FLIPPED axes
ggplot(heatmap_data, aes(x = rpa_threat_ranking, y = iucn_l2, fill = rpa_threat_ranking)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = color_ramp,
    guide = "none"
  ) +
  scale_y_discrete(limits = rev) +
  labs(
    x = "RPA Threat Ranking",
    y = "IUCN-CMP Threat (Level 2)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    text = element_text(color = "black"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 9),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

-   **Dan - the heatmap chart no longer shows a count for the number of DUs, it just lists the threat rankings available. I think that should help with the bias. I don't list the "Unknown" values in this graphic, I'm making that an assumed category. But I could add the "Unknown" values and treat missing values like Pathogens as an "Unknown"? Let me know what you think.**

## Case Study: IUCN-CMP 10.1 Geological Events

Geologic events, specifically landslides, were rated as "Extreme" for three DUs of Fraser River Chinook and one DU of Fraser sockeye (Table \@ref(tab:du-extreme-threat-table); @dfo2020; @doutaz2023). This ranking was largely due to a significant landslide on the Fraser River near Big Bar, British Columbia in late 2018 that blocked adult salmon upstream migration [@dfo2020; @doutaz2023]. The published RPAs for these populations refer to this threat category as "Avalanches and Landslides" or "Geological events (Landslides)." @dionne2023 references "IUCN-CMP threat category 10.2." However, the most current *IUCN-CMP Direct Threats Classification version 4.0* uses a slightly altered structure [@salafsky2024] where the the Level 2 threat is called 10.1 Geological Events, and avalanches and landslides are listed as examples of these events. We modified the classification of these assessments to fit the newest IUCN-CMP version for consistency.

```{r du-extreme-threat-table, message=FALSE, warning=FALSE, echo=FALSE}
library(DBI)
library(RSQLite)
library(data.table)
library(here)
library(flextable)

# Connect to the database
db_path <- file.path(here("output"), "legislation.db")
conn <- dbConnect(SQLite(), dbname = db_path)

# Query and filter the data for Extreme threat ranking only
du_data <- as.data.table(dbReadTable(conn, "cu_ranking"))
filtered_du <- du_data[rpa_threat_ranking == "Extreme", .(cu_species, du_number, du_name)]
unique_du <- unique(filtered_du)

# Disconnect
dbDisconnect(conn)

# Order by species then by DU name
unique_du <- unique_du[order(cu_species, du_name)]

# Rename columns for display
setnames(unique_du, c("cu_species", "du_number", "du_name"), c("Species", "DU Number", "Designatable Unit"))

# Create flextable for Word output
ft <- flextable(unique_du) %>%
  bold(part = "header") %>%
  align(j = c(1, 3), align = "left", part = "all") %>%
  align(j = 2, align = "center", part = "all") %>%
  set_table_properties(width = 1, layout = "autofit") %>%
  set_caption("List of Pacific salmon Designatable Units where an 'Extreme' rating was assigned to the 10.1 Geological Events IUCN-CMP threat category through the RPA process.")
ft
```

<br>

The Big Bar landslide on the Fraser River in 2018 impacted approximately 80% of the Fraser salmon stocks [@doutaz2023]. The blockage created a massive impediment to returning adult salmon, and occurred in a narrow and remote reach of the Fraser River where few options for mitigation were available. Landslides, in general, can occur naturally, but the severity and frequency can also be exacerbated by human activities such as landscape-scale changes and climate change [@doutaz2023]. This indirect connection to human activities means that 10.1 Geological Events is not a large category to be matched with regulation. The LAPSE process and tool matched eight pieces of federal and provincial legislation specific to 10.1 Geological Events (Figure \@ref(fig:fig-section-count-by-iucn)), largely related to forestry.

```{r section-count-iucn-geological, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4, fig.cap="Count of sections by legislation for IUCN threat category 10.1 Geological Events, coloured by jurisdiction."}
# Load libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = file.path(here("output"), "legislation.db"))

# Load relevant tables
paragraph_labels <- dbGetQuery(conn, "
  SELECT paragraph_id, label_value
  FROM paragraph_label_table
  WHERE label_type = 'IUCN' AND label_value = '10.1 Geological Events'
")

paragraph_table <- dbGetQuery(conn, "
  SELECT paragraph_id, legislation_id, Section
  FROM LegislationParagraphs
")

legislation_metadata <- dbGetQuery(conn, "
  SELECT legislation_id, legislation_name, jurisdiction
  FROM LegislationMetadata
")

# Disconnect from the database
dbDisconnect(conn)

# Join tables according to schema
iucn_data <- paragraph_labels %>%
  inner_join(paragraph_table, by = "paragraph_id") %>%
  inner_join(legislation_metadata, by = "legislation_id") %>%
  filter(!is.na(Section))

# Summarize unique Section-legislation_id counts by legislation and jurisdiction
summary_data <- iucn_data %>%
  mutate(section_key = paste(Section, legislation_id)) %>%
  group_by(legislation_name, jurisdiction) %>%
  summarise(Unique_sections = n_distinct(section_key), .groups = "drop")

# Reorder legislation by total section count (largest to smallest)
summary_data <- summary_data %>%
  arrange(desc(Unique_sections)) %>%
  mutate(legislation_name = factor(legislation_name, levels = rev(legislation_name)))

# Create horizontal bar chart
ggplot(summary_data, aes(
  x = Unique_sections,
  y = legislation_name,
  fill = jurisdiction
)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.8) +
  scale_fill_manual(
    values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    x = "Section Count",
    y = "Legislation",
    fill = "Jurisdiction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    text = element_text(color = "black"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

Users of the LAPSE process and tool who might be researching geological events such as landslides can get a list of possibly relevant legislation and will see that this legislation is mostly tied to forestry in British Columbia. As well, the user can see that the nature of the clauses are mostly "Instruction" and "Designation" (Figure \@ref(fig:fig-clause-type-geological-events)). The provincial *Forest Planning and Practices Regulation* under the *Forest and Range Practices Act* appears as one of the most relevant statutes specific to landslides. Examples of matched sections include: Section 3, "Damage to the environment," which states "(1)For the purpose of section 46 (1) and (1.1) *[protection of the environment]* of the Act, 'damage' means any of the following that adversely alters an ecosystem: (a)a landslide..." As well, Section 37, "Landslides," states, "An authorized person who carries out a primary forest activity must ensure that the primary forest activity does not cause a landslide that has a material adverse effect in relation to one or more of the subjects listed in section 149 (1) of the Act."

```{r fig-clause-type-geological-events, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=5, fig.cap="Section counts by clause type and legislation for IUCN-CMP threat category 10.1 Geological Events. Dot size represents section count."}
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(ggplot2)
library(stringr)

# Connect to the database
db_path <- file.path(here("output"), "legislation.db")

if (!file.exists(db_path)) {
  plot.new()
  text(0.5, 0.5, paste("Database file not found at:", db_path))
} else {
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # Get paragraph_ids for Geological Events
  geo_paragraphs <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id
    FROM paragraph_label_table
    WHERE label_type = 'IUCN' AND label_value = '10.1 Geological Events'
  "))
  
  # Get Clause Type labels for those paragraphs
  clause_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS clause_type
    FROM paragraph_label_table
    WHERE label_type = 'Clause Type' AND label_value IS NOT NULL
  "))
  
  # Get paragraph metadata
  paragraph_meta <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, Section, legislation_id
    FROM LegislationParagraphs
    WHERE Section IS NOT NULL
  "))
  
  # Get legislation names
  legislation_meta <- data.table(dbGetQuery(conn, "
    SELECT legislation_id, legislation_name
    FROM LegislationMetadata
  "))
  
  dbDisconnect(conn)
  
  # Filter clause labels to only those linked to Geological Events
  clause_geo <- merge(clause_labels, geo_paragraphs, by = "paragraph_id")
  
  # Merge with paragraph metadata
  merged <- merge(clause_geo, paragraph_meta, by = "paragraph_id", all.x = TRUE)
  
  # Merge with legislation names
  merged <- merge(merged, legislation_meta, by = "legislation_id", all.x = TRUE)
  
  # Count unique sections by clause type and legislation
  merged_summary <- merged[
    !is.na(Section),
    .(section_count = uniqueN(paste(legislation_id, Section, sep = "_"))),
    by = .(clause_type, legislation_name)
  ]
  
  # Calculate totals by clause type
  clause_totals <- merged_summary[, .(section_count = sum(section_count)), by = clause_type]
  clause_totals[, legislation_name := "TOTAL"]
  
  # Combine with main data
  merged_summary <- rbind(merged_summary, clause_totals)
  
  # Order legislation by total section count, with TOTAL at the end
  legislation_order <- c(
    merged_summary[legislation_name != "TOTAL", .(total = sum(section_count)), by = legislation_name][order(-total), legislation_name],
    "TOTAL"
  )
  merged_summary[, legislation_name := factor(legislation_name, levels = rev(legislation_order))]
  
  # Order clause types alphabetically
  clause_order <- sort(unique(merged_summary$clause_type), decreasing = FALSE)
  merged_summary[, clause_type := factor(clause_type, levels = clause_order)]
  
  # Wrap long legislation names
  merged_summary[, legislation_name_wrapped := str_wrap(legislation_name, width = 30)]
  merged_summary[, legislation_name_wrapped := factor(legislation_name_wrapped, levels = str_wrap(rev(legislation_order), width = 30))]
  
  # Create dot plot heatmap with FLIPPED AXES
  ggplot(merged_summary, aes(
    x = clause_type,
    y = legislation_name_wrapped,
    size = section_count
  )) +
    geom_point(aes(color = legislation_name == "TOTAL"), alpha = 0.8) +
    scale_size_continuous(
      range = c(2, 12),
      name = "Section Count"
    ) +
    scale_color_manual(
      values = c("FALSE" = "#76948C", "TRUE" = "#4A6A7A"),
      guide = "none"
    ) +
    geom_hline(yintercept = 1.5, linetype = "dashed", color = "grey50") +
    labs(
      x = "Clause Type",
      y = "Legislation"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      text = element_text(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
      axis.text.y = element_text(size = 8),
      axis.title = element_text(size = 11, face = "bold"),
      legend.position = "right",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 9),
      panel.grid.major = element_line(color = "#E0E0E0", linewidth = 0.3),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}
```

<br>

## Case Study: IUCN-CMP 7.2 Dams & Water Management / Use

The "Extreme-High" rating was applied to one threat category through DFO's Recovery Potential Assessment processes. The *Recovery Potential Assessment for 11 Designatable Units of Fraser River Chinook salmon, Oncorhynchus tshawytcha, Part 1: Elements 1 to 11* applied the rating to a threat category called "Natural systems modifications (Water management, ecosystems modifications)," which applies to one Fraser River Chinook DU: the South Thompson, Stream, Summer (Bessette) DU14 [\@dfo2020]. The RPA provides a rationale specific to the Bessette Creek Chinook DU:

::: {custom-style="Quote"}
Groundwater extraction is of particular concern to stream-type Chinook Salmon as these populations are highly dependent on hyporheic and groundwater inputs for much of their freshwater residence. Hyporheic and groundwater upwelling protects redds from anchor-ice formation, maintains suitable temperatures for late-summer rearing habitats, moderates temperatures and water levels for returning adults, and creates thermal refugia. Despite the critical dependence of stream-resident salmonids on hyporheic and groundwater, in the past, groundwater extraction has not been actively managed. Surface water resources are also fully subscribed in many rivers, particularly in the arid southern interior. Water extraction is of particular concern for DU14 (STh-Bessette), which spawn in a drought-sensitive system in an area of extensive water extraction. Agricultural water use creates low summer flows and high stream temperatures to the point where fish kills have been recorded. The town of Lumby is a large contributor to groundwater extraction for the surrounding area, and water demands are expected to increase in the future. Due to the deficit of water caused by anthropogenic activities this threat was deemed to have a high to extreme level of impact on this population.
:::

The *IUCN-CMP Direct Threats Classification version 4.0* lists dam construction, dam operations, levees and dikes, channelization, highway culverts, wetland draining, water catchment areas, surface water withdrawals, and groundwater pumping as examples of 7.2 Dams and Water Management / Use, and no longer lists "Natural systems modifications" as a threat category. We followed the version 4.0 category structure for consistency and considered the "Natural systems modifications" as applying to 7.2 Dams and Water Management / Use since water extraction was described as the most significant factor for this threat in the RPA.

```{r iucn-dams-water-tally, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(DBI)
library(RSQLite)
library(here)
library(data.table)

db_path <- file.path(here("output"), "legislation.db")
if (file.exists(db_path)) {
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  iucn_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS iucn_l2
    FROM paragraph_label_table
    WHERE label_type = 'IUCN' AND label_value = '7.2 Dams & Water Management / Use'
  "))
  
  paragraph_meta <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, Section, legislation_id
    FROM LegislationParagraphs
    WHERE Section IS NOT NULL
  "))
  
  dbDisconnect(conn)
  
  merged <- merge(iucn_labels, paragraph_meta, by = "paragraph_id", all.x = TRUE)
  
  iucn_tally <- merged[, .(
    clauses = uniqueN(paragraph_id),
    sections = uniqueN(paste(legislation_id, Section, sep = "_")),
    statutes = uniqueN(legislation_id)
  ), by = iucn_l2]
  
  # Store values as variables
  n_statutes <- iucn_tally$statutes[1]
  n_sections <- iucn_tally$sections[1]
  n_clauses <- iucn_tally$clauses[1]
}
```

In contrast to the previous case study, 7.2 Dams and Water Management / Use is one of the largest categories found within the accumulated legislation, matching with `r n_statutes` statutes, `r n_sections` sections, and `r n_clauses` clauses. However, just like the previous case study, the most relevant legislation is in the provincial jurisdiction (Figure \@ref(fig:fig-section-count-by-iucn)), and the BC *Water Sustainability Act* stands out as the most relevant piece of legislation.

```{r section-count-iucn-dams-water, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, fig.cap="Count of sections by legislation for IUCN threat category 7.2 Dams & Water Management / Use, coloured by jurisdiction. Top 15 legislation shown."}
# Load libraries
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(ggplot2)
library(stringr)

# Connect to the database
conn <- dbConnect(SQLite(), dbname = file.path(here("output"), "legislation.db"))

# Load relevant tables
paragraph_labels <- dbGetQuery(conn, "
  SELECT paragraph_id, label_value
  FROM paragraph_label_table
  WHERE label_type = 'IUCN' AND label_value = '7.2 Dams & Water Management / Use'
")

paragraph_table <- dbGetQuery(conn, "
  SELECT paragraph_id, legislation_id, Section
  FROM LegislationParagraphs
")

legislation_metadata <- dbGetQuery(conn, "
  SELECT legislation_id, legislation_name, jurisdiction
  FROM LegislationMetadata
")

# Disconnect from the database
dbDisconnect(conn)

# Join tables according to schema
iucn_data <- paragraph_labels %>%
  inner_join(paragraph_table, by = "paragraph_id") %>%
  inner_join(legislation_metadata, by = "legislation_id") %>%
  filter(!is.na(Section))

# Summarize unique Section-legislation_id counts by legislation and jurisdiction
summary_data <- iucn_data %>%
  mutate(section_key = paste(Section, legislation_id)) %>%
  group_by(legislation_name, jurisdiction) %>%
  summarise(Unique_sections = n_distinct(section_key), .groups = "drop")

# Get top 15 legislation by total section count
top_legislation <- summary_data %>%
  group_by(legislation_name) %>%
  summarise(total = sum(Unique_sections), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = 15) %>%
  pull(legislation_name)

# Filter to top 15
summary_data <- summary_data %>%
  filter(legislation_name %in% top_legislation)

# Wrap long legislation names (max 40 characters per line)
summary_data <- summary_data %>%
  mutate(legislation_name_wrapped = str_wrap(legislation_name, width = 40))

# Reorder by total section count (largest to smallest)
summary_data <- summary_data %>%
  arrange(desc(Unique_sections)) %>%
  mutate(legislation_name_wrapped = factor(legislation_name_wrapped, levels = rev(unique(legislation_name_wrapped))))

# Create horizontal bar chart
ggplot(summary_data, aes(
  x = Unique_sections,
  y = legislation_name_wrapped,
  fill = jurisdiction
)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3, width = 0.8) +
  scale_fill_manual(
    values = c("Federal" = "#996666", "Provincial" = "#668899", "Unknown" = "#999999"),
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    x = "Section Count",
    y = "Legislation",
    fill = "Jurisdiction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    text = element_text(color = "black"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

The clause type matches for 7.2 Dams and Water Management / Use were much more evenly distributed likely due to the large number of clause matches in total for that IUCN-CMP threat category. "Interpretation & Purpose" matched the most clauses (includes Definition sections), while "Administration, Application, & Structure" matched the fewest clauses followed by "Prohibition, Restriction, or Limitation."

```{r fig-clause-type-water-use, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=11, fig.cap="Section counts by clause type and legislation for IUCN-CMP threat category 7.2 Dams & Water Management / Use. Dot size represents section count. Top 15 legislation shown."}
library(DBI)
library(RSQLite)
library(here)
library(data.table)
library(ggplot2)
library(stringr)

# Connect to the database
db_path <- file.path(here("output"), "legislation.db")

if (!file.exists(db_path)) {
  plot.new()
  text(0.5, 0.5, paste("Database file not found at:", db_path))
} else {
  conn <- dbConnect(SQLite(), dbname = db_path)
  
  # Get paragraph_ids for Water Management / Use
  water_paragraphs <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id
    FROM paragraph_label_table
    WHERE label_type = 'IUCN' AND label_value = '7.2 Dams & Water Management / Use'
  "))
  
  # Get Clause Type labels for those paragraphs
  clause_labels <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, label_value AS clause_type
    FROM paragraph_label_table
    WHERE label_type = 'Clause Type' AND label_value IS NOT NULL
  "))
  
  # Get paragraph metadata
  paragraph_meta <- data.table(dbGetQuery(conn, "
    SELECT paragraph_id, Section, legislation_id
    FROM LegislationParagraphs
    WHERE Section IS NOT NULL
  "))
  
  # Get legislation names
  legislation_meta <- data.table(dbGetQuery(conn, "
    SELECT legislation_id, legislation_name
    FROM LegislationMetadata
  "))
  
  dbDisconnect(conn)
  
  # Filter clause labels to only those linked to Water Management / Use
  clause_water <- merge(clause_labels, water_paragraphs, by = "paragraph_id")
  
  # Merge with paragraph metadata
  merged <- merge(clause_water, paragraph_meta, by = "paragraph_id", all.x = TRUE)
  
  # Merge with legislation names
  merged <- merge(merged, legislation_meta, by = "legislation_id", all.x = TRUE)
  
  # Count unique sections by clause type and legislation
  merged_summary <- merged[
    !is.na(Section),
    .(section_count = uniqueN(paste(legislation_id, Section, sep = "_"))),
    by = .(clause_type, legislation_name)
  ]
  
  # Get top 15 legislation by total section count
  top_legislation <- merged_summary[, .(total = sum(section_count)), by = legislation_name][order(-total)][1:15, legislation_name]
  
  # Filter to top 15
  merged_summary <- merged_summary[legislation_name %in% top_legislation]
  
  # Calculate totals by clause type
  clause_totals <- merged_summary[, .(section_count = sum(section_count)), by = clause_type]
  clause_totals[, legislation_name := "TOTAL"]
  
  # Combine with main data
  merged_summary <- rbind(merged_summary, clause_totals)
  
  # Order legislation by total section count, with TOTAL at the end
  legislation_order <- c(
    merged_summary[legislation_name != "TOTAL", .(total = sum(section_count)), by = legislation_name][order(-total), legislation_name],
    "TOTAL"
  )
  merged_summary[, legislation_name := factor(legislation_name, levels = rev(legislation_order))]
  
  # Order clause types alphabetically
  clause_order <- sort(unique(merged_summary$clause_type), decreasing = FALSE)
  merged_summary[, clause_type := factor(clause_type, levels = clause_order)]
  
  # Wrap long legislation names
  merged_summary[, legislation_name_wrapped := str_wrap(legislation_name, width = 30)]
  merged_summary[, legislation_name_wrapped := factor(legislation_name_wrapped, levels = str_wrap(rev(legislation_order), width = 30))]
  
  # Create dot plot heatmap with FLIPPED AXES
  ggplot(merged_summary, aes(
    x = clause_type,
    y = legislation_name_wrapped,
    size = section_count
  )) +
    geom_point(aes(color = legislation_name == "TOTAL"), alpha = 0.8) +
    scale_size_continuous(
      range = c(2, 12),
      name = "Section Count"
    ) +
    scale_color_manual(
      values = c("FALSE" = "#76948C", "TRUE" = "#4A6A7A"),
      guide = "none"
    ) +
    geom_hline(yintercept = 1.5, linetype = "dashed", color = "grey50") +
    labs(
      x = "Clause Type",
      y = "Legislation"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      text = element_text(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
      axis.text.y = element_text(size = 8),
      axis.title = element_text(size = 11, face = "bold"),
      legend.position = "right",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 9),
      panel.grid.major = element_line(color = "#E0E0E0", linewidth = 0.3),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}
```

Users of the LAPSE process and tool researching water management and use would get a list of possibly relevant legislation, and would see that the top four matches are related to the BC *Water Sustainability Act*, including the *Groundwater Protection Regulations*. This outcome is supported by historic examples, especially in Bessette Creek watershed where the Province of British Columbia refused a groundwater use application using the *Water Sustainability Act* in 2016 citing environmental flow needs [@mccleary2021]. Both case studies show that the highest-ranking threats to Pacific salmon are governed mostly in the British Columbia provincial jurisdiction, while stock status and harvest allocations are determined federally. This relationship indicates the need for salmon stewardship to occur across collective jurisdictions.

# Discussion

Socio-political boundaries at multiple spatial scales across a species’ range increase ambiguity regarding the ownership and regulation of space [@dallimer2015]. Federal parliament and provincial legislatures rely on broadly discretionary terms in statutes that appear protective [@collins2019], but may chronically fail to prevent environmental degradation [@bankes2014]. One term can make the difference between a clause being mandatory or discretionary (“shall” as opposed to “may”), allowing an official to only exercise that authority or responsibility when economically feasible. The same laws that prohibit killing fish or damaging habitat also contain provisions to authorize those activities through licences, permits, and exemptions [@swerdfager2023]. This vagueness creates a large dependence on the executive to explain these implementation gaps [@bankes2014], allowing for arbitrary and somewhat dysfunctional stewardship styles lacking a broader overseer or responsible agency across the entire salmon life cycle.

Federal departments differ in how they implement the same legislation. For example, during the *Species At Risk Act* listing process, DFO conducts socio-economic analyses, whereas Environment and Climate Change Canada (ECCC) does not [@turcotte2021]. These misaligned processes create a situation where even when a salmon population is assessed as “endangered” through COSEWIC such as the Cultus Lake and Sakinaw Lake sockeye (@dfo2019; @ramshaw2019), the population could still be omitted from the List of Endangered Species [@governmentofcanada2005] for economic and political reasons, denying the population larger legislative protections.

# Conclusions

Pacific salmon stewardship projects (e.g. monitoring, restoration, enhancement, etc.) and decision processes often consider policies, methodologies, and literature, but the federal and provincial legislation that regulates human activities is not often reviewed. Legislative language can be vague and illusive, and the administrative structure of Pacific salmon jurisdiction in British Columbia is a decentralized and overlapping mosaic. The disconnect between legislative and natural resource language can lead to ineffective compliance with social values. The Legislation Applicable to Pacific Salmon and Ecosystems (LAPSE) process developed through this study attempts to bridge the language gap to equip practitioners with increased accessibility and navigability of the legal landscape.

The LAPSE tool provides a novel approach of linking legal language with salmon stewardship categorization by developing management domain categories based on internationally recognized biodiversity threat categories (IUCN-CMP threats). The methods developed to identify and match salmon management keywords and keyphrases to clauses allow for multiple outputs and opportunities including: an extensive list of legislation (acts, regulations, sections, and paragraphs) relevant to a human activity or salmon stewardship decisions; assessment of the nature or context of the statute (clause type) by management domain and threat; assessment of the overlap and connectedness in management domains (co-occurrence) to better understand the relationship between salmon threats in legislation. While this process and tool are relatively new and untested, further development could lead to a better integration of salmon stewardship decisions and administrative responsibility for such an important pillar of food security for communities across British Columbia.

# References

::: {#refs}
:::

# Supplementary material

-   **Links or downloads for the LAPSE tool materials (app etc.)**
